---
title: "Annotating Pellicci gamma-delta T-cell single-cell dataset (whole cell)"
description: |
author:
  - name: William Ho
    affiliation: Single Cell Open Research Endeavour (SCORE), WEHI
    affiliation_url: https://www.wehi.edu.au/people/rory-bowden/4536/wehi-advanced-genomics-facility
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
bibliography: ref.bib
---

```{r setup}
library(SingleCellExperiment)
library(here)
library(scater)
library(scran)
library(ggplot2)
library(cowplot)
library(edgeR)
library(Glimma)
library(BiocParallel)
library(patchwork)
library(janitor)
library(pheatmap)
library(batchelor)
library(rmarkdown)
library(BiocStyle)
library(readxl)
library(dplyr)
library(tidyr)
library(ggrepel)
library(magrittr)

knitr::opts_chunk$set(fig.path = "C094_Pellicci.single-cell.annotate.whole_cell_files/")
```

# Preparing the data

We start from the cell selected *SingleCellExperiment* object created in ['Merging cells for Pellicci gamma-delta T-cell dataset (whole cell)'](C094_Pellicci.single-cell.merge.whole_cell.html).

```{r}
sce <- readRDS(here("data", "SCEs", "C094_Pellicci.single-cell.merge.whole_cell.SCE.rds"))

# pre-create directories for saving export, or error (dir not exists)
dir.create(here("data", "marker_genes", "whole_cell"), recursive = TRUE)
dir.create(here("output", "marker_genes", "whole_cell"), recursive = TRUE)

# remove "Unknown" (as it is not informative at all)
sce <- sce[, sce$stage != "Unknown"]
colData(sce) <- droplevels(colData(sce))

# Some useful colours
plate_number_colours <- setNames(
  unique(sce$colours$plate_number_colours),
  unique(names(sce$colours$plate_number_colours)))
plate_number_colours <- plate_number_colours[levels(sce$plate_number)]
tissue_colours <- setNames(
  unique(sce$colours$tissue_colours),
  unique(names(sce$colours$tissue_colours)))
tissue_colours <- tissue_colours[levels(sce$tissue)]
donor_colours <- setNames(
  unique(sce$colours$donor_colours),
  unique(names(sce$colours$donor_colours)))
donor_colours <- donor_colours[levels(sce$donor)]
stage_colours <- setNames(
  unique(sce$colours$stage_colours),
  unique(names(sce$colours$stage_colours)))
stage_colours <- stage_colours[levels(sce$stage)]
group_colours <- setNames(
  unique(sce$colours$group_colours),
  unique(names(sce$colours$group_colours)))
group_colours <- group_colours[levels(sce$group)]
cluster_colours <- setNames(
  unique(sce$colours$cluster_colours),
  unique(names(sce$colours$cluster_colours)))
cluster_colours <- cluster_colours[levels(sce$cluster)]

# Some useful gene sets
mito_set <- rownames(sce)[any(rowData(sce)$ENSEMBL.SEQNAME == "MT")]
ribo_set <- grep("^RP(S|L)", rownames(sce), value = TRUE)
# NOTE: A more curated approach for identifying ribosomal protein genes
#       (https://github.com/Bioconductor/OrchestratingSingleCellAnalysis-base/blob/ae201bf26e3e4fa82d9165d8abf4f4dc4b8e5a68/feature-selection.Rmd#L376-L380)
library(msigdbr)
c2_sets <- msigdbr(species = "Homo sapiens", category = "C2")
ribo_set <- union(
  ribo_set,
  c2_sets[c2_sets$gs_name == "KEGG_RIBOSOME", ]$gene_symbol)
ribo_set <- intersect(ribo_set, rownames(sce))
sex_set <- rownames(sce)[any(rowData(sce)$ENSEMBL.SEQNAME %in% c("X", "Y"))]
pseudogene_set <- rownames(sce)[
  any(grepl("pseudogene", rowData(sce)$ENSEMBL.GENEBIOTYPE))]
# NOTE: not suggest to narrow down into protein coding genes (pcg) as it remove all significant candidate in most of the comparison !!!
protein_coding_gene_set <- rownames(sce)[
  any(grepl("protein_coding", rowData(sce)$ENSEMBL.GENEBIOTYPE))]

# # gene-of-interest (suggested by Dan at meeting on 15 Jul 2021)
# interest <- c("KLRB1", "CD4", # QC
#              "NKG7", "GNLY", "CCL5", # cluster 1 selected unique markers
#              "TCF7", "SOX4", "LEF1", "CD3D", "CCR9", "BCL11B", "NREP" # cluster 3 selected unique markers
# )
# NOTE: to prevent bias and confusion of the gene above are DE to all cases, I decide to include only the QC genes, i.e. "CD4" and "KLBR1", in each heatmap
```

# Data overview

```{r clusterplot-umap, fig.cap = "UMAP plot, where each point represents a cell and is coloured according to the legend.", fig.asp = 1.25}
p1 <- plotReducedDim(sce, "UMAP_corrected", colour_by = "cluster", theme_size = 7, point_size = 0.2) +
  scale_colour_manual(values = cluster_colours, name = "cluster")
p2 <- plotReducedDim(sce, "UMAP_corrected", colour_by = "stage", theme_size = 7, point_size = 0.2) +
  scale_colour_manual(values = stage_colours, name = "stage")
p3 <- plotReducedDim(sce, "UMAP_corrected", colour_by = "plate_number", theme_size = 7, point_size = 0.2) +
  scale_colour_manual(values = plate_number_colours, name = "plate_number")
p4 <- plotReducedDim(sce, "UMAP_corrected", colour_by = "tissue", theme_size = 7, point_size = 0.2) +
  scale_colour_manual(values = tissue_colours, name = "tissue")
p5 <- plotReducedDim(sce, "UMAP_corrected", colour_by = "donor", theme_size = 7, point_size = 0.2) +
  scale_colour_manual(values = donor_colours, name = "donor")
p6 <- plotReducedDim(sce, "UMAP_corrected", colour_by = "group", theme_size = 7, point_size = 0.2) +
  scale_colour_manual(values = group_colours, name = "group")
(p1 | p2) / (p3 | p4) / (p5 | p6)
```

```{r cluster-barplot, fig.asp = 1, fig.cap = "Breakdown of clusters by experimental factors."}
# summary - stacked barplot
p1 <- ggcells(sce) +
  geom_bar(aes(x = cluster, fill = cluster)) +
  coord_flip() +
  ylab("Number of samples") +
  theme_cowplot(font_size = 8) +
  scale_fill_manual(values = cluster_colours) +
  geom_text(stat='count', aes(x = cluster, label=..count..), hjust=1.5, size=2)
p2 <- ggcells(sce) +
  geom_bar(
    aes(x = cluster, fill = stage),
    position = position_fill(reverse = TRUE)) +
  coord_flip() +
  ylab("Frequency") +
  scale_fill_manual(values = stage_colours) +
  theme_cowplot(font_size = 8)
p3 <- ggcells(sce) +
  geom_bar(
    aes(x = cluster, fill = plate_number),
    position = position_fill(reverse = TRUE)) +
  coord_flip() +
  ylab("Frequency") +
  scale_fill_manual(values = plate_number_colours) +
  theme_cowplot(font_size = 8)
p4 <- ggcells(sce) +
  geom_bar(
    aes(x = cluster, fill = tissue),
    position = position_fill(reverse = TRUE)) +
  coord_flip() +
  ylab("Frequency") +
  scale_fill_manual(values = tissue_colours) +
  theme_cowplot(font_size = 8)
p5 <- ggcells(sce) +
  geom_bar(
    aes(x = cluster, fill = donor),
    position = position_fill(reverse = TRUE)) +
  coord_flip() +
  ylab("Frequency") +
  scale_fill_manual(values = donor_colours) +
  theme_cowplot(font_size = 8)
p6 <- ggcells(sce) +
  geom_bar(
    aes(x = cluster, fill = group),
    position = position_fill(reverse = TRUE)) +
  coord_flip() +
  ylab("Frequency") +
  scale_fill_manual(values = group_colours) +
  theme_cowplot(font_size = 8)
(p1 | p2) / (p3 | p4) / (p5 | p6)
```

**NOTE**: Considering the fact that SingleR with use of the annotation reference (Monaco Immune Cell Data) most relevant to the gamma-delta T cells (even annotated at `cell` level) could not further sub-classify the developmental stage/subtype of them (cluster `4`: `Th1 cell`-/`Naive CD8`/`CD4 T cell`-alike; all other clusters, i.e. `1`, `2`, `3`: `Vd2gd T cells`) [if interested, refer to: EDA_annotation_SingleR_MI_fine_cell_level.R], we decide to characterize the clusters by manual detection and curation of specific marker genes directly.

# Marker gene detection

To interpret our clustering results, we identify the genes that drive separation between clusters.
These marker genes allow us to assign biological meaning to each cluster based on their functional annotation.
In the most obvious case, the marker genes for each cluster are *a priori* associated with particular cell types, allowing us to treat the clustering as a proxy for cell type identity.
The same principle can be applied to more subtle differences in activation status or differentiation state.

Identification of marker genes is usually based around the retrospective detection of differential expression between clusters^[Differential expression analyses is performed on the original log-expression values.
We do not use the MNN-corrected values here except to obtain the clusters to be characterized.].
Genes that are more strongly DE are more likely to have driven cluster separation in the first place.
The top DE genes are likely to be good candidate markers as they can effectively distinguish between cells in different clusters.

## Using pairwise *t*-tests

The Welch *t*-test is an obvious choice of statistical method to test for differences in expression between clusters.
It is quickly computed and has good statistical properties for large numbers of cells [@soneson2018bias].

```{r}
# block on plate
sce$block <- paste0(sce$plate_number)
```

## Cluster `1` vs. `2` vs. `3` vs. `4`

Here we look for the unique up-regulated markers of each cluster when compared to the all remaining ones.
For instance, unique markers of cluster `1` refer to the genes significantly up-regulated in **all** of these comparisons: cluster `2` vs. `1` **and** cluster `3` vs. `1` **and** cluster `4` vs. `1`.

```{r}
###################################
# cluster 1 (i.e. S3-mix more Blood) vs 2 (i.e. S3-mix more Thymus) vs 3 (i.e. mostly Thymus.S1 + .S2) vs 4 (i.e. mostly Thymus.S3)

# find unique DE ./. clusters
uniquely_up <- findMarkers(
  sce,
  groups = sce$cluster,
  block = sce$block,
  pval.type = "all",
  direction = "up")
```

```{r}
# export DGE lists
saveRDS(
  uniquely_up,
  here("data", "marker_genes", "whole_cell", "C094_Pellicci.uniquely_up.cluster_1_vs_2_vs_3_vs_4.rds"),
  compress = "xz")

dir.create(here("output", "marker_genes", "whole_cell", "uniquely_up", "cluster_1_vs_2_vs_3_vs_4"), recursive = TRUE)

vs_pair <- c("1", "2", "3", "4")

message("Writing 'uniquely_up (cluster_1_vs_2_vs_3_vs_4)' marker genes to file.")
for (n in names(uniquely_up)) {
  message(n)
  gzout <- gzfile(
    description = here(
      "output",
      "marker_genes",
      "whole_cell",
      "uniquely_up",
      "cluster_1_vs_2_vs_3_vs_4",
      paste0("cluster_",
             vs_pair[which(names(uniquely_up) %in% n)],
             "_vs_",
             vs_pair[-which(names(uniquely_up) %in% n)][1],
             "_vs_",
             vs_pair[-which(names(uniquely_up) %in% n)][2],
             "_vs_",
             vs_pair[-which(names(uniquely_up) %in% n)][3],
             ".uniquely_up.csv.gz")),
    open = "wb")
  write.table(
    x = uniquely_up[[n]] %>%
      as.data.frame() %>%
      tibble::rownames_to_column("gene_ID"),
    file = gzout,
    sep = ",",
    quote = FALSE,
    row.names = FALSE,
    col.names = TRUE)
  close(gzout)
}
```

```{r, echo = FALSE}
# NOTE: The following is a workaround to the lack of support for tabsets in 
#       distill (see https://github.com/rstudio/distill/issues/11 and 
#       https://github.com/rstudio/distill/issues/11#issuecomment-692142414 in 
#       particular).
xaringanExtra::use_panelset()
```

::::: {.panelset}

::: {.panel}

#### `Cluster 1` {.panel-name}

```{r}
##########################################
# look at cluster 1 (i.e. S3-mix more Blood)
chosen <- "1"
cluster1_uniquely_up <- uniquely_up[[chosen]]

# add description for the chosen cluster-group
x <- "(S3-mix more Blood)"

# look only at protein coding gene (pcg)
# NOTE: not suggest to narrow down into pcg as it remove all significant candidates (FDR << 0.05) !
# cluster1_uniquely_up <- cluster1_uniquely_up[intersect(protein_coding_gene_set, rownames(cluster1_uniquely_up)), ]

# get rid of noise (i.e. pseudo, ribo, mito, sex) that collaborator not interested in
cluster1_uniquely_up_noiseR <- cluster1_uniquely_up[setdiff(rownames(cluster1_uniquely_up), c(pseudogene_set, mito_set, ribo_set, sex_set)), ]

# see if key marker, "CD4 and/or ""KLRB1/CD161"", contain in the DE list + if it is "significant (i.e FDR <0.05)
y <- c("CD4",
       which(rownames(cluster1_uniquely_up_noiseR) %in% "CD4"),
       cluster1_uniquely_up_noiseR[which(rownames(cluster1_uniquely_up_noiseR) %in% "CD4"), ]$FDR < 0.05)
z <- c("KLRB1/CD161",
       which(rownames(cluster1_uniquely_up_noiseR) %in% "KLRB1"),
       cluster1_uniquely_up_noiseR[which(rownames(cluster1_uniquely_up_noiseR) %in% "KLRB1"), ]$FDR < 0.05)

# top25 only + gene-of-interest
best_set <- cluster1_uniquely_up_noiseR[1:25, ]
```

```{r heat-uniquely-up-logExp-cluster-1, fig.cap = "Heatmap of log-expression values in each sample for the top uniquely upregulated marker genes. Each column is a sample, each row a gene. Colours are capped at -3 and 3 to preserve dynamic range. Ranking of CD4 and CD161/KLRB1 from top of the DGE list sorted in ascending order of FDR and their statistical significance (TRUE = FDR < 0.05) are provided in the title", fig.asp = 1.2, layout="l-body-outset"}
# heatmap
plotHeatmap(
  sce,
  features = rownames(best_set),
  columns = order(
    sce$cluster,
    sce$stage,
    sce$tissue,
    sce$donor,
    sce$group,
    sce$plate_number),
  colour_columns_by = c(
    "cluster",
    "stage",
    "tissue",
    "donor",
    "group",
    "plate_number"),
  cluster_cols = FALSE,
  center = TRUE,
  symmetric = TRUE,
  zlim = c(-3, 3),
  show_colnames = FALSE,
  annotation_row = data.frame(
    Sig = factor(
      ifelse(best_set[, "FDR"] < 0.05, "Yes", "No"),
      # TODO: temp trick to deal with the row-colouring problem
      # levels = c("Yes", "No")),
      levels = c("Yes")),
    row.names = rownames(best_set)),
  main = paste0("Cluster: ", chosen, " ", x, " - \n",
                y[1], "_top ", y[2], "_significance: ", y[3], " ; \n",
                z[1], "_top ", z[2], "_significance: ", z[3]),
  column_annotation_colors = list(
    # Sig = c("Yes" = "red", "No" = "lightgrey"),
    cluster = cluster_colours,
    stage = stage_colours,
    tissue = tissue_colours,
    donor = donor_colours,
    group = group_colours,
    plate_number = plate_number_colours),
  color = hcl.colors(101, "Blue-Red 3"),
  fontsize = 7)
```

:::

::: {.panel}

#### `Cluster 2` {.panel-name}

```{r}
##########################################
# look at cluster 2 (i.e. S3-mix more Thymus)
chosen <- "2"
cluster2_uniquely_up <- uniquely_up[[chosen]]

# add description for the chosen cluster-group
x <- "(S3-mix more Thymus)"

# look only at protein coding gene (pcg)
# NOTE: not suggest to narrow down into pcg as it remove all significant candidates (FDR << 0.05) !
# cluster2_uniquely_up <- cluster2_uniquely_up[intersect(protein_coding_gene_set, rownames(cluster2_uniquely_up)), ]

# get rid of noise (i.e. pseudo, ribo, mito, sex) that collaborator not interested in
cluster2_uniquely_up_noiseR <- cluster2_uniquely_up[setdiff(rownames(cluster2_uniquely_up), c(pseudogene_set, mito_set, ribo_set, sex_set)), ]

# see if key marker, "CD4 and/or ""KLRB1/CD161"", contain in the DE list + if it is "significant (i.e FDR <0.05)
y <- c("CD4",
       which(rownames(cluster2_uniquely_up_noiseR) %in% "CD4"),
       cluster2_uniquely_up_noiseR[which(rownames(cluster2_uniquely_up_noiseR) %in% "CD4"), ]$FDR < 0.05)
z <- c("KLRB1/CD161",
       which(rownames(cluster2_uniquely_up_noiseR) %in% "KLRB1"),
       cluster2_uniquely_up_noiseR[which(rownames(cluster2_uniquely_up_noiseR) %in% "KLRB1"), ]$FDR < 0.05)

# top25 only + gene-of-interest
best_set <- cluster2_uniquely_up_noiseR[1:25, ]
```

```{r heat-uniquely-up-logExp-cluster-2, fig.cap = "Heatmap of log-expression values in each sample for the top uniquely upregulated marker genes. Each column is a sample, each row a gene. Colours are capped at -3 and 3 to preserve dynamic range. Ranking of CD4 and CD161/KLRB1 from top of the DGE list sorted in ascending order of FDR and their statistical significance (TRUE = FDR < 0.05) are provided in the title.", fig.asp = 1.2, layout="l-body-outset"}
# heatmap
plotHeatmap(
  sce,
  features = rownames(best_set),
  columns = order(
    sce$cluster,
    sce$stage,
    sce$tissue,
    sce$donor,
    sce$group,
    sce$plate_number),
  colour_columns_by = c(
    "cluster",
    "stage",
    "tissue",
    "donor",
    "group",
    "plate_number"),
  cluster_cols = FALSE,
  center = TRUE,
  symmetric = TRUE,
  zlim = c(-3, 3),
  show_colnames = FALSE,
  # annotation_row = data.frame(
  #   Sig = factor(
  #     ifelse(best_set[, "FDR"] < 0.05, "Yes", "No"),
  #     # TODO: temp trick to deal with the row-colouring problem
  #     # levels = c("Yes", "No")),
  #     levels = c("No")),
  #   row.names = rownames(best_set)),
  main = paste0("Cluster: ", chosen, " ", x, " - \n",
                y[1], "_top ", y[2], "_significance: ", y[3], " ; \n",
                z[1], "_top ", z[2], "_significance: ", z[3]),
  column_annotation_colors = list(
    # Sig = c("Yes" = "red", "No" = "lightgrey"),
    cluster = cluster_colours,
    stage = stage_colours,
    tissue = tissue_colours,
    donor = donor_colours,
    group = group_colours,
    plate_number = plate_number_colours),
  color = hcl.colors(101, "Blue-Red 3"),
  fontsize = 7)
```

:::

::: {.panel}

#### `Cluster 3` {.panel-name}

```{r}
##########################################
# look at cluster 3 (i.e. mostly Thymus.S1 + .S2)
chosen <- "3"
cluster3_uniquely_up <- uniquely_up[[chosen]]

# add description for the chosen cluster-group
x <- "(mostly Thymus.S1 + .S2)"

# look only at protein coding gene (pcg)
# NOTE: not suggest to narrow down into pcg as it remove all significant candidates (FDR << 0.05) !
# cluster3_uniquely_up <- cluster3_uniquely_up[intersect(protein_coding_gene_set, rownames(cluster3_uniquely_up)), ]

# get rid of noise (i.e. pseudo, ribo, mito, sex) that collaborator not interested in
cluster3_uniquely_up_noiseR <- cluster3_uniquely_up[setdiff(rownames(cluster3_uniquely_up), c(pseudogene_set, mito_set, ribo_set, sex_set)), ]

# see if key marker, "CD4 and/or ""KLRB1/CD161"", contain in the DE list + if it is "significant (i.e FDR <0.05)
y <- c("CD4",
       which(rownames(cluster3_uniquely_up_noiseR) %in% "CD4"),
       cluster3_uniquely_up_noiseR[which(rownames(cluster3_uniquely_up_noiseR) %in% "CD4"), ]$FDR < 0.05)
z <- c("KLRB1/CD161",
       which(rownames(cluster3_uniquely_up_noiseR) %in% "KLRB1"),
       cluster3_uniquely_up_noiseR[which(rownames(cluster3_uniquely_up_noiseR) %in% "KLRB1"), ]$FDR < 0.05)

# top25 only + gene-of-interest
best_set <- cluster3_uniquely_up_noiseR[1:25, ]
```

```{r heat-uniquely-up-logExp-cluster-3, fig.cap = "Heatmap of log-expression values in each sample for the top uniquely upregulated marker genes. Each column is a sample, each row a gene. Colours are capped at -3 and 3 to preserve dynamic range. Ranking of CD4 and CD161/KLRB1 from top of the DGE list sorted in ascending order of FDR and their statistical significance (TRUE = FDR < 0.05) are provided in the title.", fig.asp = 1.2, layout="l-body-outset"}
# heatmap
plotHeatmap(
  sce,
  features = rownames(best_set),
  columns = order(
    sce$cluster,
    sce$stage,
    sce$tissue,
    sce$donor,
    sce$group,
    sce$plate_number),
  colour_columns_by = c(
    "cluster",
    "stage",
    "tissue",
    "donor",
    "group",
    "plate_number"),
  cluster_cols = FALSE,
  center = TRUE,
  symmetric = TRUE,
  zlim = c(-3, 3),
  show_colnames = FALSE,
  annotation_row = data.frame(
    Sig = factor(
      ifelse(best_set[, "FDR"] < 0.05, "Yes", "No"),
      # TODO: temp trick to deal with the row-colouring problem
      # levels = c("Yes", "No")),
      levels = c("Yes")),
    row.names = rownames(best_set)),
  main = paste0("Cluster: ", chosen, " ", x, " - \n",
                y[1], "_top ", y[2], "_significance: ", y[3], " ; \n",
                z[1], "_top ", z[2], "_significance: ", z[3]),
  column_annotation_colors = list(
    # Sig = c("Yes" = "red", "No" = "lightgrey"),
    cluster = cluster_colours,
    stage = stage_colours,
    tissue = tissue_colours,
    donor = donor_colours,
    group = group_colours,
    plate_number = plate_number_colours),
  color = hcl.colors(101, "Blue-Red 3"),
  fontsize = 7)
```

:::

::: {.panel}

#### `Cluster 4` {.panel-name}

```{r}
##########################################
# look at cluster 4 (i.e. mostly Thymus.S3)
chosen <- "4"
cluster4_uniquely_up <- uniquely_up[[chosen]]

# add description for the chosen cluster-group
x <- "(mostly Thymus.S3)"

# look only at protein coding gene (pcg)
# NOTE: not suggest to narrow down into pcg as it remove all significant candidates (FDR << 0.05) !
# cluster4_uniquely_up <- cluster4_uniquely_up[intersect(protein_coding_gene_set, rownames(cluster4_uniquely_up)), ]

# get rid of noise (i.e. pseudo, ribo, mito, sex) that collaborator not interested in
cluster4_uniquely_up_noiseR <- cluster4_uniquely_up[setdiff(rownames(cluster4_uniquely_up), c(pseudogene_set, mito_set, ribo_set, sex_set)), ]

# see if key marker, "CD4 and/or ""KLRB1/CD161"", contain in the DE list + if it is "significant (i.e FDR <0.05)
y <- c("CD4",
       which(rownames(cluster4_uniquely_up_noiseR) %in% "CD4"),
       cluster4_uniquely_up_noiseR[which(rownames(cluster4_uniquely_up_noiseR) %in% "CD4"), ]$FDR < 0.05)
z <- c("KLRB1/CD161",
       which(rownames(cluster4_uniquely_up_noiseR) %in% "KLRB1"),
       cluster4_uniquely_up_noiseR[which(rownames(cluster4_uniquely_up_noiseR) %in% "KLRB1"), ]$FDR < 0.05)

# top25 only + gene-of-interest
best_set <- cluster4_uniquely_up_noiseR[1:25, ]
```

```{r heat-uniquely-up-logExp-cluster-4, fig.cap = "Heatmap of log-expression values in each sample for the top uniquely upregulated marker genes. Each column is a sample, each row a gene. Colours are capped at -3 and 3 to preserve dynamic range. Ranking of CD4 and CD161/KLRB1 from top of the DGE list sorted in ascending order of FDR and their statistical significance (TRUE = FDR < 0.05) are provided in the title.", fig.asp = 1.2, layout="l-body-outset"}
# heatmap
plotHeatmap(
  sce,
  features = rownames(best_set),
  columns = order(
    sce$cluster,
    sce$stage,
    sce$tissue,
    sce$donor,
    sce$group,
    sce$plate_number),
  colour_columns_by = c(
    "cluster",
    "stage",
    "tissue",
    "donor",
    "group",
    "plate_number"),
  cluster_cols = FALSE,
  center = TRUE,
  symmetric = TRUE,
  zlim = c(-3, 3),
  show_colnames = FALSE,
  annotation_row = data.frame(
    Sig = factor(
      ifelse(best_set[, "FDR"] < 0.05, "Yes", "No"),
      # TODO: temp trick to deal with the row-colouring problem
      # levels = c("Yes", "No")),
      levels = c("Yes")),
    row.names = rownames(best_set)),
  main = paste0("Cluster: ", chosen, " ", x, " - \n",
                y[1], "_top ", y[2], "_significance: ", y[3], " ; \n",
                z[1], "_top ", z[2], "_significance: ", z[3]),
  column_annotation_colors = list(
    # Sig = c("Yes" = "red", "No" = "lightgrey"),
    cluster = cluster_colours,
    stage = stage_colours,
    tissue = tissue_colours,
    donor = donor_colours,
    group = group_colours,
    plate_number = plate_number_colours),
  color = hcl.colors(101, "Blue-Red 3"),
  fontsize = 7)
```

:::

::::::

DGE lists of these comparisons are available in [`output/marker_genes/whole_cell/uniquely_up/cluster_1_vs_2_vs_3_vs_4/`](..output/marker_genes/whole_cell/uniquely_up/cluster_1_vs_2_vs_3_vs_4/).

**Summary**: Stemmed from the fact that the cluster `1`, `2`, and `4` are too similar to each other, except for cluster `3`, we cannot determine sufficient number of marker genes for characterizing each of these clusters.

With this regard, apart from making "all pairwise comparisons" between "all clusters" to pinpoint DE unique to each cluster as above, we took an alternative path and determined the DE unique to only the "selected pairwise comparisons" between "clusters".
Say, for cluster `3`, we determine markers that is significantly up-regulated in **at least one** of these comparisons: cluster `1` vs. `3` **or** cluster `2` vs. `3` **or** cluster `4` vs. `3`. 

Besides, we also look into the the pairwise comparisons between the interesting "cluster-groups".
For instance, cluster `3` is the only cluster relatively distant from the main groups in the UMAP (Figure \@ref(fig:clusterplot-umap)), thus it would be interesting to know what drive cluster `3` from the merged group of cluster `1`, `2`, and `4`, or maybe cluster `1` and `2`.

## Selected pairwise comparisons

```{r, echo = FALSE}
# NOTE: The following is a workaround to the lack of support for tabsets in 
#       distill (see https://github.com/rstudio/distill/issues/11 and 
#       https://github.com/rstudio/distill/issues/11#issuecomment-692142414 in 
#       particular).
xaringanExtra::use_panelset()
```

::::: {.panelset}

::: {.panel}

#### `cluster_1+2+4_vs_cluster_3` {.panel-name}

```{r}
#########
# A vs B
#########

##########################################################################
# cluster 1 + 2 + 4 (i.e. mostly S3) vs cluster 3 (i.e. mostly S1 and S2)

# checkpoint
cp <- sce

# classify cluster-group for comparison
cp$vs1 <- factor(ifelse(cp$cluster == 1 | cp$cluster == 2 | cp$cluster == 4, "A", "B"))

# set vs colours
vs1_colours <- setNames(
  palette.colors(nlevels(cp$vs1), "Set1"),
  levels(cp$vs1))
cp$colours$vs1_colours <- vs1_colours[cp$vs1]

# find unique DE ./. cluster-groups
vs1_uniquely_up <- findMarkers(
  cp,
  groups = cp$vs1,
  block = cp$block,
  pval.type = "all",
  direction = "up")

# export DGE lists
saveRDS(
  vs1_uniquely_up,
  here("data", "marker_genes", "whole_cell", "C094_Pellicci.uniquely_up.cluster_1_2_4_vs_3.rds"),
  compress = "xz")

dir.create(here("output", "marker_genes", "whole_cell", "uniquely_up", "cluster_1_2_4_vs_3"), recursive = TRUE)

vs_pair <- c("1_2_4", "3")

message("Writing 'uniquely_up (cluster_1_2_4_vs_3)' marker genes to file.")
for (n in names(vs1_uniquely_up)) {
  message(n)
  gzout <- gzfile(
    description = here(
      "output",
      "marker_genes",
      "whole_cell",
      "uniquely_up",
      "cluster_1_2_4_vs_3",
      paste0("cluster_",
             vs_pair[which(names(vs1_uniquely_up) %in% n)],
             "_vs_",
             vs_pair[-which(names(vs1_uniquely_up) %in% n)][1],
             ".uniquely_up.csv.gz")),
    open = "wb")
  write.table(
    x = vs1_uniquely_up[[n]] %>%
      as.data.frame() %>%
      tibble::rownames_to_column("gene_ID"),
    file = gzout,
    sep = ",",
    quote = FALSE,
    row.names = FALSE,
    col.names = TRUE)
  close(gzout)
}
```

```{r}
##########################################################
# look at cluster-group A / cluster 1+2+4 (i.e. mostly S3)
chosen <- "A"
A_uniquely_up <- vs1_uniquely_up[[chosen]]

# add description for the chosen cluster-group
x <- "(cluster 1+2+4; mostly S3)"

# look only at protein coding gene (pcg)
# NOTE: not suggest to narrow down into pcg as it remove all significant candidates (FDR << 0.05) !
# A_uniquely_up_pcg <- A_uniquely_up[intersect(protein_coding_gene_set, rownames(A_uniquely_up)), ]

# get rid of noise (i.e. pseudo, ribo, mito, sex) that collaborator not interested in
A_uniquely_up_noiseR <- A_uniquely_up[setdiff(rownames(A_uniquely_up), c(pseudogene_set, mito_set, ribo_set, sex_set)), ]

# see if key marker, "CD4 and/or ""KLRB1/CD161"", contain in the DE list + if it is "significant (i.e FDR <0.05)
y <- c("CD4",
       which(rownames(A_uniquely_up_noiseR) %in% "CD4"),
       A_uniquely_up_noiseR[which(rownames(A_uniquely_up_noiseR) %in% "CD4"), ]$FDR < 0.05)
z <- c("KLRB1/CD161",
       which(rownames(A_uniquely_up_noiseR) %in% "KLRB1"),
       A_uniquely_up_noiseR[which(rownames(A_uniquely_up_noiseR) %in% "KLRB1"), ]$FDR < 0.05)

# top25 only + gene-of-interest
best_set <- A_uniquely_up_noiseR[1:25, ]
```

```{r heat-uniquely-up-logExp-cluster-1-2-4-vs-3, fig.cap = "Heatmap of log-expression values in each sample for the top uniquely upregulated marker genes. Each column is a sample, each row a gene. Colours are capped at -3 and 3 to preserve dynamic range. Ranking of CD4 and CD161/KLRB1 from top of the DGE list sorted in ascending order of FDR and their statistical significance (TRUE = FDR < 0.05) are provided in the title.", fig.asp = 1.2, layout="l-body-outset"}
# heatmap
plotHeatmap(
  cp,
  features = rownames(best_set),
  columns = order(
    cp$vs1,
    cp$cluster,
    cp$stage,
    cp$tissue,
    cp$donor,
    cp$group,
    cp$plate_number),
  colour_columns_by = c(
    "vs1",
    "cluster",
    "stage",
    "tissue",
    "donor",
    "group",
    "plate_number"),
  cluster_cols = FALSE,
  center = TRUE,
  symmetric = TRUE,
  zlim = c(-3, 3),
  show_colnames = FALSE,
  annotation_row = data.frame(
    Sig = factor(
      ifelse(best_set[, "FDR"] < 0.05, "Yes", "No"),
      # TODO: temp trick to deal with the row-colouring problem
      # levels = c("Yes", "No")),
      levels = c("Yes")),
    row.names = rownames(best_set)),
  main = paste0("Cluster-group: ", chosen, " ", x, " - \n",
                y[1], "_top ", y[2], "_significance: ", y[3], " ; \n",
                z[1], "_top ", z[2], "_significance: ", z[3]),
  column_annotation_colors = list(
    # Sig = c("Yes" = "red", "No" = "lightgrey"),
    vs1 = vs1_colours,
    cluster = cluster_colours,
    stage = stage_colours,
    tissue = tissue_colours,
    donor = donor_colours,
    group = group_colours,
    plate_number = plate_number_colours),
  color = hcl.colors(101, "Blue-Red 3"),
  fontsize = 7)
```

```{r}
###########################################################
# look at cluster-group B / cluster 3 (i.e. mostly S1 + S2)
chosen <- "B"
B_uniquely_up <- vs1_uniquely_up[[chosen]]

# add description for the chosen cluster-group
x <- "(cluster 3; mostly S1 and S2)"

# look only at protein coding gene (pcg)
# NOTE: not suggest to narrow down into pcg as it remove all significant candidates (FDR << 0.05) !
# B_uniquely_up_pcg <- B_uniquely_up[intersect(protein_coding_gene_set, rownames(B_uniquely_up)), ]

# get rid of noise (i.e. pseudo, ribo, mito, sex) that collaborator not interested in
B_uniquely_up_noiseR <- B_uniquely_up[setdiff(rownames(B_uniquely_up), c(pseudogene_set, mito_set, ribo_set, sex_set)), ]

# see if key marker, "CD4 and/or ""KLRB1/CD161"", contain in the DE list + if it is "significant (i.e FDR <0.05)
y <- c("CD4",
       which(rownames(B_uniquely_up_noiseR) %in% "CD4"),
       B_uniquely_up_noiseR[which(rownames(B_uniquely_up_noiseR) %in% "CD4"), ]$FDR < 0.05)
z <- c("KLRB1/CD161",
       which(rownames(B_uniquely_up_noiseR) %in% "KLRB1"),
       B_uniquely_up_noiseR[which(rownames(B_uniquely_up_noiseR) %in% "KLRB1"), ]$FDR < 0.05)

# top25 only + gene-of-interest
best_set <- B_uniquely_up_noiseR[1:25, ]
```

```{r heat-uniquely-up-logExp-cluster-3-vs-1-2-4, fig.cap = "Heatmap of log-expression values in each sample for the top uniquely upregulated marker genes. Each column is a sample, each row a gene. Colours are capped at -3 and 3 to preserve dynamic range. Ranking of CD4 and CD161/KLRB1 from top of the DGE list sorted in ascending order of FDR and their statistical significance (TRUE = FDR < 0.05) are provided in the title.", fig.asp = 1.2, layout="l-body-outset"}
# heatmap
plotHeatmap(
  cp,
  features = rownames(best_set),
  columns = order(
    cp$vs1,
    cp$cluster,
    cp$stage,
    cp$tissue,
    cp$donor,
    cp$group,
    cp$plate_number),
  colour_columns_by = c(
    "vs1",
    "cluster",
    "stage",
    "tissue",
    "donor",
    "group",
    "plate_number"),
  cluster_cols = FALSE,
  center = TRUE,
  symmetric = TRUE,
  zlim = c(-3, 3),
  show_colnames = FALSE,
  annotation_row = data.frame(
    Sig = factor(
      ifelse(best_set[, "FDR"] < 0.05, "Yes", "No"),
      # TODO: temp trick to deal with the row-colouring problem
      # levels = c("Yes", "No")),
      levels = c("Yes")),
    row.names = rownames(best_set)),
  main = paste0("Cluster-group: ", chosen, " ", x, " - \n",
                y[1], "_top ", y[2], "_significance: ", y[3], " ; \n",
                z[1], "_top ", z[2], "_significance: ", z[3]),
  column_annotation_colors = list(
    # Sig = c("Yes" = "red", "No" = "lightgrey"),
    vs1 = vs1_colours,
    cluster = cluster_colours,
    stage = stage_colours,
    tissue = tissue_colours,
    donor = donor_colours,
    group = group_colours,
    plate_number = plate_number_colours),
  color = hcl.colors(101, "Blue-Red 3"),
  fontsize = 7)
```

DGE lists of these comparisons are available in [`output/marker_genes/whole_cell/uniquely_up/cluster_1_2_4_vs_3/`](..output/marker_genes/whole_cell/uniquely_up/cluster_1_2_4_vs_3/).

:::

::: {.panel}

#### `cluster_1+2_vs_cluster_3` {.panel-name}

```{r}
#########
# C vs D
#########

##########################################################################################
# cluster 1 + 2 (i.e. mostly Blood.S3) vs cluster 3 (i.e. mostly Thymus.S1 and Thymus.S2)

# checkpoint
cp <- sce
# exclude cells of uninterested cluster from cp
cp <- cp[, cp$cluster != "4"]
colData(cp) <- droplevels(colData(cp))

# classify cluster-group for comparison
cp$vs2 <- factor(ifelse(cp$cluster == 1 | cp$cluster == 2 , "C", "D"))

# set vs colours
vs2_colours <- setNames(
  palette.colors(nlevels(cp$vs2), "Set1"),
  levels(cp$vs2))
cp$colours$vs2_colours <- vs2_colours[cp$vs2]

# find unique DE ./. cluster-groups
vs2_uniquely_up <- findMarkers(
  cp,
  groups = cp$vs2,
  block = cp$block,
  pval.type = "all",
  direction = "up")

# export DGE lists
saveRDS(
  vs2_uniquely_up,
  here("data", "marker_genes", "whole_cell", "C094_Pellicci.uniquely_up.cluster_1_2_vs_3.rds"),
  compress = "xz")

dir.create(here("output", "marker_genes", "whole_cell", "uniquely_up", "cluster_1_2_vs_3"), recursive = TRUE)

vs_pair <- c("1_2", "3")

message("Writing 'uniquely_up (cluster_1_2_vs_3)' marker genes to file.")
for (n in names(vs2_uniquely_up)) {
  message(n)
  gzout <- gzfile(
    description = here(
      "output",
      "marker_genes",
      "whole_cell",
      "uniquely_up",
      "cluster_1_2_vs_3",
      paste0("cluster_",
             vs_pair[which(names(vs2_uniquely_up) %in% n)],
             "_vs_",
             vs_pair[-which(names(vs2_uniquely_up) %in% n)][1],
             ".uniquely_up.csv.gz")),
    open = "wb")
  write.table(
    x = vs2_uniquely_up[[n]] %>%
      as.data.frame() %>%
      tibble::rownames_to_column("gene_ID"),
    file = gzout,
    sep = ",",
    quote = FALSE,
    row.names = FALSE,
    col.names = TRUE)
  close(gzout)
}
```

```{r}
###############################################################
# look at cluster-group C / cluster 1+2 (i.e. mostly Blood.S3)
chosen <- "C"
C_uniquely_up <- vs2_uniquely_up[[chosen]]

# add description for the chosen cluster-group
x <- "(cluster 1+2; mostly Blood.S3)"

# look only at protein coding gene (pcg)
# NOTE: not suggest to narrow down into pcg as it remove all significant candidates (FDR << 0.05) !
# C_uniquely_up_pcg <- C_uniquely_up[intersect(protein_coding_gene_set, rownames(C_uniquely_up)), ]

# get rid of noise (i.e. pseudo, ribo, mito, sex) that collaborator not interested in
C_uniquely_up_noiseR <- C_uniquely_up[setdiff(rownames(C_uniquely_up), c(pseudogene_set, mito_set, ribo_set, sex_set)), ]

# see if key marker, "CD4 and/or ""KLRB1/CD161"", contain in the DE list + if it is "significant (i.e FDR <0.05)
y <- c("CD4",
       which(rownames(C_uniquely_up_noiseR) %in% "CD4"),
       C_uniquely_up_noiseR[which(rownames(C_uniquely_up_noiseR) %in% "CD4"), ]$FDR < 0.05)
z <- c("KLRB1/CD161",
       which(rownames(C_uniquely_up_noiseR) %in% "KLRB1"),
       C_uniquely_up_noiseR[which(rownames(C_uniquely_up_noiseR) %in% "KLRB1"), ]$FDR < 0.05)

# top25 only + gene-of-interest
best_set <- C_uniquely_up_noiseR[1:25, ]
```

```{r heat-uniquely-up-logExp-cluster-1-2-vs-3, fig.cap = "Heatmap of log-expression values in each sample for the top uniquely upregulated marker genes. Each column is a sample, each row a gene. Colours are capped at -3 and 3 to preserve dynamic range. Ranking of CD4 and CD161/KLRB1 from top of the DGE list sorted in ascending order of FDR and their statistical significance (TRUE = FDR < 0.05) are provided in the title.", fig.asp = 1.2, layout="l-body-outset"}
# heatmap
plotHeatmap(
  cp,
  features = rownames(best_set),
  columns = order(
    cp$vs2,
    cp$cluster,
    cp$stage,
    cp$tissue,
    cp$donor,
    cp$group,
    cp$plate_number),
  colour_columns_by = c(
    "vs2",
    "cluster",
    "stage",
    "tissue",
    "donor",
    "group",
    "plate_number"),
  cluster_cols = FALSE,
  center = TRUE,
  symmetric = TRUE,
  zlim = c(-3, 3),
  show_colnames = FALSE,
  annotation_row = data.frame(
    Sig = factor(
      ifelse(best_set[, "FDR"] < 0.05, "Yes", "No"),
      # TODO: temp trick to deal with the row-colouring problem
      # levels = c("Yes", "No")),
      levels = c("Yes")),
    row.names = rownames(best_set)),
  main = paste0("Cluster-group: ", chosen, " ", x, " - \n",
                y[1], "_top ", y[2], "_significance: ", y[3], " ; \n",
                z[1], "_top ", z[2], "_significance: ", z[3]),
  column_annotation_colors = list(
    # Sig = c("Yes" = "red", "No" = "lightgrey"),
    vs2 = vs2_colours,
    cluster = cluster_colours,
    stage = stage_colours,
    tissue = tissue_colours,
    donor = donor_colours,
    group = group_colours,
    plate_number = plate_number_colours),
  color = hcl.colors(101, "Blue-Red 3"),
  fontsize = 7)
```

```{r}
###########################################################################
# look at cluster-group D / cluster 3 (i.e. mostly Thymus.S1 and Thymus.S2)
chosen <- "D"
D_uniquely_up <- vs2_uniquely_up[[chosen]]

# add description for the chosen cluster-group
x <- "(cluster 3; mostly Thymus.S1 and Thymus.S2)"

# look only at protein coding gene (pcg)
# NOTE: not suggest to narrow down into pcg as it remove all significant candidates (FDR << 0.05) !
# D_uniquely_up_pcg <- D_uniquely_up[intersect(protein_coding_gene_set, rownames(D_uniquely_up)), ]

# get rid of noise (i.e. pseudo, ribo, mito, sex) that collaborator not interested in
D_uniquely_up_noiseR <- D_uniquely_up[setdiff(rownames(D_uniquely_up), c(pseudogene_set, mito_set, ribo_set, sex_set)), ]

# see if key marker, "CD4 and/or ""KLRB1/CD161"", contain in the DE list + if it is "significant (i.e FDR <0.05)
y <- c("CD4",
       which(rownames(D_uniquely_up_noiseR) %in% "CD4"),
       D_uniquely_up_noiseR[which(rownames(D_uniquely_up_noiseR) %in% "CD4"), ]$FDR < 0.05)
z <- c("KLRB1/CD161",
       which(rownames(D_uniquely_up_noiseR) %in% "KLRB1"),
       D_uniquely_up_noiseR[which(rownames(D_uniquely_up_noiseR) %in% "KLRB1"), ]$FDR < 0.05)

# top25 only + gene-of-interest
best_set <- D_uniquely_up_noiseR[1:25, ]
```

```{r heat-uniquely-up-logExp-cluster-3-vs-1-2, fig.cap = "Heatmap of log-expression values in each sample for the top uniquely upregulated marker genes. Each column is a sample, each row a gene. Colours are capped at -3 and 3 to preserve dynamic range. Ranking of CD4 and CD161/KLRB1 from top of the DGE list sorted in ascending order of FDR and their statistical significance (TRUE = FDR < 0.05) are provided in the title.", fig.asp = 1.2, layout="l-body-outset"}
# heatmap
plotHeatmap(
  cp,
  features = rownames(best_set),
  columns = order(
    cp$vs2,
    cp$cluster,
    cp$stage,
    cp$tissue,
    cp$donor,
    cp$group,
    cp$plate_number),
  colour_columns_by = c(
    "vs2",
    "cluster",
    "stage",
    "tissue",
    "donor",
    "group",
    "plate_number"),
  cluster_cols = FALSE,
  center = TRUE,
  symmetric = TRUE,
  zlim = c(-3, 3),
  show_colnames = FALSE,
  annotation_row = data.frame(
    Sig = factor(
      ifelse(best_set[, "FDR"] < 0.05, "Yes", "No"),
      # TODO: temp trick to deal with the row-colouring problem
      # levels = c("Yes", "No")),
      levels = c("Yes")),
    row.names = rownames(best_set)),
  main = paste0("Cluster-group: ", chosen, " ", x, " - \n",
                y[1], "_top ", y[2], "_significance: ", y[3], " ; \n",
                z[1], "_top ", z[2], "_significance: ", z[3]),
  column_annotation_colors = list(
    # Sig = c("Yes" = "red", "No" = "lightgrey"),
    vs2 = vs2_colours,
    cluster = cluster_colours,
    stage = stage_colours,
    tissue = tissue_colours,
    donor = donor_colours,
    group = group_colours,
    plate_number = plate_number_colours),
  color = hcl.colors(101, "Blue-Red 3"),
  fontsize = 7)
```

DGE lists of these comparisons are available in [`output/marker_genes/whole_cell/uniquely_up/cluster_1_2_vs_3/`](..output/marker_genes/whole_cell/uniquely_up/cluster_1_2_vs_3/).

:::

::: {.panel}

#### `cluster_1+2_vs_cluster_4` {.panel-name}

```{r}
#########
# E vs F
#########

##########################################################################
# cluster 1 + 2 (i.e. mostly Blood.S3) vs cluster 4 (i.e. mostly Thymus.S3)

# checkpoint
cp <- sce
# exclude cells of uninterested cluster from cp
cp <- cp[, cp$cluster != "3"]
colData(cp) <- droplevels(colData(cp))
# classify cluster-group for comparison

cp$vs3 <- factor(ifelse(cp$cluster == 1 | cp$cluster == 2 , "E", "F"))

# set vs colours
vs3_colours <- setNames(
  palette.colors(nlevels(cp$vs3), "Set1"),
  levels(cp$vs3))
cp$colours$vs3_colours <- vs3_colours[cp$vs3]

# find unique DE ./. cluster-groups
vs3_uniquely_up <- findMarkers(
  cp,
  groups = cp$vs3,
  block = cp$block,
  pval.type = "all",
  direction = "up")

# export DGE lists
saveRDS(
  vs3_uniquely_up,
  here("data", "marker_genes", "whole_cell", "C094_Pellicci.uniquely_up.cluster_1_2_vs_4.rds"),
  compress = "xz")

dir.create(here("output", "marker_genes", "whole_cell", "uniquely_up", "cluster_1_2_vs_4"), recursive = TRUE)

vs_pair <- c("1_2", "4")

message("Writing 'uniquely_up (cluster_1_2_vs_4)' marker genes to file.")
for (n in names(vs3_uniquely_up)) {
  message(n)
  gzout <- gzfile(
    description = here(
      "output",
      "marker_genes",
      "whole_cell",
      "uniquely_up",
      "cluster_1_2_vs_4",
      paste0("cluster_",
             vs_pair[which(names(vs3_uniquely_up) %in% n)],
             "_vs_",
             vs_pair[-which(names(vs3_uniquely_up) %in% n)][1],
             ".uniquely_up.csv.gz")),
    open = "wb")
  write.table(
    x = vs3_uniquely_up[[n]] %>%
      as.data.frame() %>%
      tibble::rownames_to_column("gene_ID"),
    file = gzout,
    sep = ",",
    quote = FALSE,
    row.names = FALSE,
    col.names = TRUE)
  close(gzout)
}
```

```{r}
###############################################################
# look at cluster-group E / cluster 1+2 (i.e. mostly Blood.S3)
chosen <- "E"
E_uniquely_up <- vs3_uniquely_up[[chosen]]

# add description for the chosen cluster-group
x <- "(cluster 1+2; mostly Blood.S3)"

# look only at protein coding gene (pcg)
# NOTE: not suggest to narrow down into pcg as it remove all significant candidates (FDR << 0.05) !
# E_uniquely_up_pcg <- E_uniquely_up[intersect(protein_coding_gene_set, rownames(E_uniquely_up)), ]

# get rid of noise (i.e. pseudo, ribo, mito, sex) that collaborator not interested in
E_uniquely_up_noiseR <- E_uniquely_up[setdiff(rownames(E_uniquely_up), c(pseudogene_set, mito_set, ribo_set, sex_set)), ]

# see if key marker, "CD4 and/or ""KLRB1/CD161"", contain in the DE list + if it is "significant (i.e FDR <0.05)
y <- c("CD4",
       which(rownames(E_uniquely_up_noiseR) %in% "CD4"),
       E_uniquely_up_noiseR[which(rownames(E_uniquely_up_noiseR) %in% "CD4"), ]$FDR < 0.05)
z <- c("KLRB1/CD161",
       which(rownames(E_uniquely_up_noiseR) %in% "KLRB1"),
       E_uniquely_up_noiseR[which(rownames(E_uniquely_up_noiseR) %in% "KLRB1"), ]$FDR < 0.05)

# top25 only + gene-of-interest
best_set <- E_uniquely_up_noiseR[1:25, ]
```

```{r heat-uniquely-up-logExp-cluster-1-2-vs-4, fig.cap = "Heatmap of log-expression values in each sample for the top uniquely upregulated marker genes. Each column is a sample, each row a gene. Colours are capped at -3 and 3 to preserve dynamic range. Ranking of CD4 and CD161/KLRB1 from top of the DGE list sorted in ascending order of FDR and their statistical significance (TRUE = FDR < 0.05) are provided in the title.", fig.asp = 1.2, layout="l-body-outset"}
# heatmap
plotHeatmap(
  cp,
  features = rownames(best_set),
  columns = order(
    cp$vs3,
    cp$cluster,
    cp$stage,
    cp$tissue,
    cp$donor,
    cp$group,
    cp$plate_number),
  colour_columns_by = c(
    "vs3",
    "cluster",
    "stage",
    "tissue",
    "donor",
    "group",
    "plate_number"),
  cluster_cols = FALSE,
  center = TRUE,
  symmetric = TRUE,
  zlim = c(-3, 3),
  show_colnames = FALSE,
  annotation_row = data.frame(
    Sig = factor(
      ifelse(best_set[, "FDR"] < 0.05, "Yes", "No"),
      # TODO: temp trick to deal with the row-colouring problem
      # levels = c("Yes", "No")),
      levels = c("Yes")),
    row.names = rownames(best_set)),
  main = paste0("Cluster-group: ", chosen, " ", x, " - \n",
                y[1], "_top ", y[2], "_significance: ", y[3], " ; \n",
                z[1], "_top ", z[2], "_significance: ", z[3]),
  column_annotation_colors = list(
    # Sig = c("Yes" = "red", "No" = "lightgrey"),
    vs3 = vs3_colours,
    cluster = cluster_colours,
    stage = stage_colours,
    tissue = tissue_colours,
    donor = donor_colours,
    group = group_colours,
    plate_number = plate_number_colours),
  color = hcl.colors(101, "Blue-Red 3"),
  fontsize = 7)
```

```{r}
##############################################################
# look at cluster-group F / cluster 4 (i.e. mostly Thymus.S3)
chosen <- "F"
F_uniquely_up <- vs3_uniquely_up[[chosen]]

# add description for the chosen cluster-group
x <- "(cluster 4; mostly Thymus.S3)"

# look only at protein coding gene (pcg)
# NOTE: not suggest to narrow down into pcg as it remove all significant candidates (FDR << 0.05) !
# F_uniquely_up_pcg <- F_uniquely_up[intersect(protein_coding_gene_set, rownames(F_uniquely_up)), ]

# get rid of noise (i.e. pseudo, ribo, mito, sex) that collaborator not interested in
F_uniquely_up_noiseR <- F_uniquely_up[setdiff(rownames(F_uniquely_up), c(pseudogene_set, mito_set, ribo_set, sex_set)), ]

# see if key marker, "CD4 and/or ""KLRB1/CD161"", contain in the DE list + if it is "significant (i.e FDR <0.05)
y <- c("CD4",
       which(rownames(F_uniquely_up_noiseR) %in% "CD4"),
       F_uniquely_up_noiseR[which(rownames(F_uniquely_up_noiseR) %in% "CD4"), ]$FDR < 0.05)
z <- c("KLRB1/CD161",
       which(rownames(F_uniquely_up_noiseR) %in% "KLRB1"),
       F_uniquely_up_noiseR[which(rownames(F_uniquely_up_noiseR) %in% "KLRB1"), ]$FDR < 0.05)

# top25 only + gene-of-interest
best_set <- F_uniquely_up_noiseR[1:25, ]
```

```{r heat-uniquely-up-logExp-cluster-4-vs-1-2, fig.cap = "Heatmap of log-expression values in each sample for the top uniquely upregulated marker genes. Each column is a sample, each row a gene. Colours are capped at -3 and 3 to preserve dynamic range. Ranking of CD4 and CD161/KLRB1 from top of the DGE list sorted in ascending order of FDR and their statistical significance (TRUE = FDR < 0.05) are provided in the title.", fig.asp = 1.2, layout="l-body-outset"}
# heatmap
plotHeatmap(
  cp,
  features = rownames(best_set),
  columns = order(
    cp$vs3,
    cp$cluster,
    cp$stage,
    cp$tissue,
    cp$donor,
    cp$group,
    cp$plate_number),
  colour_columns_by = c(
    "vs3",
    "cluster",
    "stage",
    "tissue",
    "donor",
    "group",
    "plate_number"),
  cluster_cols = FALSE,
  center = TRUE,
  symmetric = TRUE,
  zlim = c(-3, 3),
  show_colnames = FALSE,
  annotation_row = data.frame(
    Sig = factor(
      ifelse(best_set[, "FDR"] < 0.05, "Yes", "No"),
      # TODO: temp trick to deal with the row-colouring problem
      # levels = c("Yes", "No")),
      levels = c("Yes")),
    row.names = rownames(best_set)),
  main = paste0("Cluster-group: ", chosen, " ", x, " - \n",
                y[1], "_top ", y[2], "_significance: ", y[3], " ; \n",
                z[1], "_top ", z[2], "_significance: ", z[3]),
  column_annotation_colors = list(
    # Sig = c("Yes" = "red", "No" = "lightgrey"),
    vs3 = vs3_colours,
    cluster = cluster_colours,
    stage = stage_colours,
    tissue = tissue_colours,
    donor = donor_colours,
    group = group_colours,
    plate_number = plate_number_colours),
  color = hcl.colors(101, "Blue-Red 3"),
  fontsize = 7)
```

DGE lists of these comparisons are available in [`output/marker_genes/whole_cell/uniquely_up/cluster_1_2_vs_4/`](..output/marker_genes/whole_cell/uniquely_up/cluster_1_2_vs_4/).

:::

::: {.panel}

#### `cluster_1_vs_cluster_2` {.panel-name}

```{r}
#########
# G vs H
#########

##########################################################################
# cluster 1 (i.e. S3-mix more Blood) vs cluster 2 (i.e. S3-mix more Thymus)

# checkpoint
cp <- sce
# exclude cells of uninterested cluster from cp
cp <- cp[, cp$cluster == "1" | cp$cluster == "2"]
colData(cp) <- droplevels(colData(cp))

# classify cluster-group for comparison
cp$vs4 <- factor(ifelse(cp$cluster == 1 , "G", "H"))

# set vs colours
vs4_colours <- setNames(
  palette.colors(nlevels(cp$vs4), "Set1"),
  levels(cp$vs4))
cp$colours$vs4_colours <- vs4_colours[cp$vs4]

# find unique DE ./. cluster-groups
vs4_uniquely_up <- findMarkers(
  cp,
  groups = cp$vs4,
  block = cp$block,
  pval.type = "all",
  direction = "up")

# export DGE lists
saveRDS(
  vs4_uniquely_up,
  here("data", "marker_genes", "whole_cell", "C094_Pellicci.uniquely_up.cluster_1_vs_2.rds"),
  compress = "xz")

dir.create(here("output", "marker_genes", "whole_cell", "uniquely_up", "cluster_1_vs_2"), recursive = TRUE)

vs_pair <- c("1", "2")

message("Writing 'uniquely_up (cluster_1_vs_2)' marker genes to file.")
for (n in names(vs4_uniquely_up)) {
  message(n)
  gzout <- gzfile(
    description = here(
      "output",
      "marker_genes",
      "whole_cell",
      "uniquely_up",
      "cluster_1_vs_2",
      paste0("cluster_",
             vs_pair[which(names(vs4_uniquely_up) %in% n)],
             "_vs_",
             vs_pair[-which(names(vs4_uniquely_up) %in% n)][1],
             ".uniquely_up.csv.gz")),
    open = "wb")
  write.table(
    x = vs4_uniquely_up[[n]] %>%
      as.data.frame() %>%
      tibble::rownames_to_column("gene_ID"),
    file = gzout,
    sep = ",",
    quote = FALSE,
    row.names = FALSE,
    col.names = TRUE)
  close(gzout)
}
```

```{r}
###########################################################
# look at cluster-group G / cluster 1 (i.e. S3-mix more Blood)
chosen <- "G"
G_uniquely_up <- vs4_uniquely_up[[chosen]]

# add description for the chosen cluster-group
x <- "(cluster 1; S3-mix more Blood)"

# look only at protein coding gene (pcg)
# NOTE: not suggest to narrow down into pcg as it remove all significant candidates (FDR << 0.05) !
# G_uniquely_up_pcg <- G_uniquely_up[intersect(protein_coding_genG_set, rownames(G_uniquely_up)), ]

# get rid of noise (i.e. pseudo, ribo, mito, sex) that collaborator not interested in
G_uniquely_up_noiseR <- G_uniquely_up[setdiff(rownames(G_uniquely_up), c(pseudogene_set, mito_set, ribo_set, sex_set)), ]

# see if key marker, "CD4 and/or ""KLRB1/CD161"", contain in the DE list + if it is "significant (i.e FDR <0.05)
y <- c("CD4",
       which(rownames(G_uniquely_up_noiseR) %in% "CD4"),
       G_uniquely_up_noiseR[which(rownames(G_uniquely_up_noiseR) %in% "CD4"), ]$FDR < 0.05)
z <- c("KLRB1/CD161",
       which(rownames(G_uniquely_up_noiseR) %in% "KLRB1"),
       G_uniquely_up_noiseR[which(rownames(G_uniquely_up_noiseR) %in% "KLRB1"), ]$FDR < 0.05)

# top25 only + gene-of-interest
best_set <- G_uniquely_up_noiseR[1:25, ]
```

```{r heat-uniquely-up-logExp-cluster-1-vs-2, fig.cap = "Heatmap of log-expression values in each sample for the top uniquely upregulated marker genes. Each column is a sample, each row a gene. Colours are capped at -3 and 3 to preserve dynamic range. Ranking of CD4 and CD161/KLRB1 from top of the DGE list sorted in ascending order of FDR and their statistical significance (TRUE = FDR < 0.05) are provided in the title.", fig.asp = 1.2, layout="l-body-outset"}
# heatmap
plotHeatmap(
  cp,
  features = rownames(best_set),
  columns = order(
    cp$vs4,
    cp$cluster,
    cp$stage,
    cp$tissue,
    cp$donor,
    cp$group,
    cp$plate_number),
  colour_columns_by = c(
    "vs4",
    "cluster",
    "stage",
    "tissue",
    "donor",
    "group",
    "plate_number"),
  cluster_cols = FALSE,
  center = TRUE,
  symmetric = TRUE,
  zlim = c(-3, 3),
  show_colnames = FALSE,
  annotation_row = data.frame(
    Sig = factor(
      ifelse(best_set[, "FDR"] < 0.05, "Yes", "No"),
      # TODO: temp trick to deal with the row-colouring problem
      # levels = c("Yes", "No")),
      levels = c("Yes")),
    row.names = rownames(best_set)),
  main = paste0("Cluster-group: ", chosen, " ", x, " - \n",
                y[1], "_top ", y[2], "_significance: ", y[3], " ; \n",
                z[1], "_top ", z[2], "_significance: ", z[3]),
  column_annotation_colors = list(
    # Sig = c("Yes" = "red", "No" = "lightgrey"),
    vs4 = vs4_colours,
    cluster = cluster_colours,
    stage = stage_colours,
    tissue = tissue_colours,
    donor = donor_colours,
    group = group_colours,
    plate_number = plate_number_colours),
  color = hcl.colors(101, "Blue-Red 3"),
  fontsize = 7)
```

```{r}
############################################################
# look at cluster-group H / cluster 2 (i.e. S3-mix more Thymus)
chosen <- "H"
H_uniquely_up <- vs4_uniquely_up[[chosen]]

# add description for the chosen cluster-group
x <- "(cluster 2; mostly Thymus.S3)"

# look only at protein coding gene (pcg)
# NOTE: not suggest to narrow down into pcg as it remove all significant candidates (FDR << 0.05) !
# H_uniquely_up_pcg <- H_uniquely_up[intersect(protein_coding_genG_set, rownames(H_uniquely_up)), ]

# get rid of noise (i.e. pseudo, ribo, mito, sex) that collaborator not interested in
H_uniquely_up_noiseR <- H_uniquely_up[setdiff(rownames(H_uniquely_up), c(pseudogene_set, mito_set, ribo_set, sex_set)), ]

# see if key marker, "CD4 and/or ""KLRB1/CD161"", contain in the DE list + if it is "significant (i.e FDR <0.05)
y <- c("CD4",
       which(rownames(H_uniquely_up_noiseR) %in% "CD4"),
       H_uniquely_up_noiseR[which(rownames(H_uniquely_up_noiseR) %in% "CD4"), ]$FDR < 0.05)
z <- c("KLRB1/CD161",
       which(rownames(H_uniquely_up_noiseR) %in% "KLRB1"),
       H_uniquely_up_noiseR[which(rownames(H_uniquely_up_noiseR) %in% "KLRB1"), ]$FDR < 0.05)

# top25 only + gene-of-interest
best_set <- H_uniquely_up_noiseR[1:25, ]
```

```{r heat-uniquely-up-logExp-cluster-2-vs-1, fig.cap = "Heatmap of log-expression values in each sample for the top uniquely upregulated marker genes. Each column is a sample, each row a gene. Colours are capped at -3 and 3 to preserve dynamic range. Ranking of CD4 and CD161/KLRB1 from top of the DGE list sorted in ascending order of FDR and their statistical significance (TRUE = FDR < 0.05) are provided in the title.", fig.asp = 1.2, layout="l-body-outset"}
# heatmap
plotHeatmap(
  cp,
  features = rownames(best_set),
  columns = order(
    cp$vs4,
    cp$cluster,
    cp$stage,
    cp$tissue,
    cp$donor,
    cp$group,
    cp$plate_number),
  colour_columns_by = c(
    "vs4",
    "cluster",
    "stage",
    "tissue",
    "donor",
    "group",
    "plate_number"),
  cluster_cols = FALSE,
  center = TRUE,
  symmetric = TRUE,
  zlim = c(-3, 3),
  show_colnames = FALSE,
  # annotation_row = data.frame(
  #   Sig = factor(
  #     ifelse(best_set[, "FDR"] < 0.05, "Yes", "No"),
  #     # TODO: temp trick to deal with the row-colouring problem
  #     # levels = c("Yes", "No")),
  #     levels = c("No")),
  #   row.names = rownames(best_set)),
  main = paste0("Cluster-group: ", chosen, " ", x, " - \n",
                y[1], "_top ", y[2], "_significance: ", y[3], " ; \n",
                z[1], "_top ", z[2], "_significance: ", z[3]),
  column_annotation_colors = list(
    # Sig = c("Yes" = "red", "No" = "lightgrey"),
    vs4 = vs4_colours,
    cluster = cluster_colours,
    stage = stage_colours,
    tissue = tissue_colours,
    donor = donor_colours,
    group = group_colours,
    plate_number = plate_number_colours),
  color = hcl.colors(101, "Blue-Red 3"),
  fontsize = 7)
```

DGE lists of these comparisons are available in [`output/marker_genes/whole_cell/uniquely_up/cluster_1_vs_2/`](..output/marker_genes/whole_cell/uniquely_up/cluster_2_vs_1/).

:::

::: {.panel}

#### `cluster_3_vs_cluster_4` {.panel-name}

```{r}
#########
# I vs J
#########

#################################################################################
# cluster 3 (i.e. mostly Thymus.S1 and .S2) vs cluster 4 (i.e. mostly Thymus.S3)

# checkpoint
cp <- sce
# exclude cells of uninterested cluster from cp
cp <- cp[, cp$cluster == "3" | cp$cluster == "4"]
colData(cp) <- droplevels(colData(cp))

# classify cluster-group for comparison
cp$vs5 <- factor(ifelse(cp$cluster == 3 , "I", "J"))

# set vs colours
vs5_colours <- setNames(
  palette.colors(nlevels(cp$vs5), "Set1"),
  levels(cp$vs5))
cp$colours$vs5_colours <- vs5_colours[cp$vs5]

# find unique DE ./. cluster-groups
vs5_uniquely_up <- findMarkers(
  cp,
  groups = cp$vs5,
  block = cp$block,
  pval.type = "all",
  direction = "up")

# export DGE lists
saveRDS(
  vs5_uniquely_up,
  here("data", "marker_genes", "whole_cell", "C094_Pellicci.uniquely_up.cluster_3_vs_4.rds"),
  compress = "xz")

dir.create(here("output", "marker_genes", "whole_cell", "uniquely_up", "cluster_3_vs_4"), recursive = TRUE)

vs_pair <- c("3", "4")

message("Writing 'uniquely_up (cluster_3_vs_4)' marker genes to file.")
for (n in names(vs5_uniquely_up)) {
  message(n)
  gzout <- gzfile(
    description = here(
      "output",
      "marker_genes",
      "whole_cell",
      "uniquely_up",
      "cluster_3_vs_4",
      paste0("cluster_",
             vs_pair[which(names(vs5_uniquely_up) %in% n)],
             "_vs_",
             vs_pair[-which(names(vs5_uniquely_up) %in% n)][1],
             ".uniquely_up.csv.gz")),
    open = "wb")
  write.table(
    x = vs5_uniquely_up[[n]] %>%
      as.data.frame() %>%
      tibble::rownames_to_column("gene_ID"),
    file = gzout,
    sep = ",",
    quote = FALSE,
    row.names = FALSE,
    col.names = TRUE)
  close(gzout)
}
```

```{r}
#####################################################################
# look at cluster-group I / cluster 3 (i.e. mostly Thymus.S1 and .S2)
chosen <- "I"
I_uniquely_up <- vs5_uniquely_up[[chosen]]

# add description for the chosen cluster-group
x <- "(cluster 3; mostly Thymus.S1 and .S2)"

# look only at protein coding gene (pcg)
# NOTE: not suggest to narrow down into pcg as it remove all significant candidates (FDR << 0.05) !
# I_uniquely_up_pcg <- I_uniquely_up[intersect(protein_coding_genI_set, rownames(I_uniquely_up)), ]

# get rid of noise (i.e. pseudo, ribo, mito, sex) that collaborator not interested in
I_uniquely_up_noiseR <- I_uniquely_up[setdiff(rownames(I_uniquely_up), c(pseudogene_set, mito_set, ribo_set, sex_set)), ]

# see if key marker, "CD4 and/or ""KLRB1/CD161"", contain in the DE list + if it is "significant (i.e FDR <0.05)
y <- c("CD4",
       which(rownames(I_uniquely_up_noiseR) %in% "CD4"),
       I_uniquely_up_noiseR[which(rownames(I_uniquely_up_noiseR) %in% "CD4"), ]$FDR < 0.05)
z <- c("KLRB1/CD161",
       which(rownames(I_uniquely_up_noiseR) %in% "KLRB1"),
       I_uniquely_up_noiseR[which(rownames(I_uniquely_up_noiseR) %in% "KLRB1"), ]$FDR < 0.05)

# top25 only + gene-of-interest
best_set <- I_uniquely_up_noiseR[1:25, ]
```

```{r heat-uniquely-up-logExp-cluster-3-vs-4, fig.cap = "Heatmap of log-expression values in each sample for the top uniquely upregulated marker genes. Each column is a sample, each row a gene. Colours are capped at -3 and 3 to preserve dynamic range. Ranking of CD4 and CD161/KLRB1 from top of the DGE list sorted in ascending order of FDR and their statistical significance (TRUE = FDR < 0.05) are provided in the title.", fig.asp = 1.2, layout="l-body-outset"}
# heatmap
plotHeatmap(
  cp,
  features = rownames(best_set),
  columns = order(
    cp$vs5,
    cp$cluster,
    cp$stage,
    cp$tissue,
    cp$donor,
    cp$group,
    cp$plate_number),
  colour_columns_by = c(
    "vs5",
    "cluster",
    "stage",
    "tissue",
    "donor",
    "group",
    "plate_number"),
  cluster_cols = FALSE,
  center = TRUE,
  symmetric = TRUE,
  zlim = c(-3, 3),
  show_colnames = FALSE,
  annotation_row = data.frame(
    Sig = factor(
      ifelse(best_set[, "FDR"] < 0.05, "Yes", "No"),
      # TODO: temp trick to deal with the row-colouring problem
      # levels = c("Yes", "No")),
      levels = c("Yes")),
    row.names = rownames(best_set)),
  main = paste0("Cluster-group: ", chosen, " ", x, " - \n",
                y[1], "_top ", y[2], "_significance: ", y[3], " ; \n",
                z[1], "_top ", z[2], "_significance: ", z[3]),
  column_annotation_colors = list(
    # Sig = c("Yes" = "red", "No" = "lightgrey"),
    vs5 = vs5_colours,
    cluster = cluster_colours,
    stage = stage_colours,
    tissue = tissue_colours,
    donor = donor_colours,
    group = group_colours,
    plate_number = plate_number_colours),
  color = hcl.colors(101, "Blue-Red 3"),
  fontsize = 7)
```

```{r}
##############################################################
# look at cluster-group J / cluster 4  (i.e. mostly Thymus.S3)
chosen <- "J"
J_uniquely_up <- vs5_uniquely_up[[chosen]]

# add description for the chosen cluster-group
x <- "(cluster 4; mostly Thymus.S3)"

# look only at protein coding gene (pcg)
# NOTE: not suggest to narrow down into pcg as it remove all significant candidates (FDR << 0.05) !
# J_uniquely_up_pcg <- J_uniquely_up[intersect(protein_coding_genI_set, rownames(J_uniquely_up)), ]

# get rid of noise (i.e. pseudo, ribo, mito, sex) that collaborator not interested in
J_uniquely_up_noiseR <- J_uniquely_up[setdiff(rownames(J_uniquely_up), c(pseudogene_set, mito_set, ribo_set, sex_set)), ]

# see if key marker, "CD4 and/or ""KLRB1/CD161"", contain in the DE list + if it is "significant (i.e FDR <0.05)
y <- c("CD4",
       which(rownames(J_uniquely_up_noiseR) %in% "CD4"),
       J_uniquely_up_noiseR[which(rownames(J_uniquely_up_noiseR) %in% "CD4"), ]$FDR < 0.05)
z <- c("KLRB1/CD161",
       which(rownames(J_uniquely_up_noiseR) %in% "KLRB1"),
       J_uniquely_up_noiseR[which(rownames(J_uniquely_up_noiseR) %in% "KLRB1"), ]$FDR < 0.05)

# top25 only + gene-of-interest
best_set <- J_uniquely_up_noiseR[1:25, ]
```

```{r heat-uniquely-up-logExp-cluster-4-vs-3, fig.cap = "Heatmap of log-expression values in each sample for the top uniquely upregulated marker genes. Each column is a sample, each row a gene. Colours are capped at -3 and 3 to preserve dynamic range. Ranking of CD4 and CD161/KLRB1 from top of the DGE list sorted in ascending order of FDR and their statistical significance (TRUE = FDR < 0.05) are provided in the title.", fig.asp = 1.2, layout="l-body-outset"}
# heatmap
plotHeatmap(
  cp,
  features = rownames(best_set),
  columns = order(
    cp$vs5,
    cp$cluster,
    cp$stage,
    cp$tissue,
    cp$donor,
    cp$group,
    cp$plate_number),
  colour_columns_by = c(
    "vs5",
    "cluster",
    "stage",
    "tissue",
    "donor",
    "group",
    "plate_number"),
  cluster_cols = FALSE,
  center = TRUE,
  symmetric = TRUE,
  zlim = c(-3, 3),
  show_colnames = FALSE,
  annotation_row = data.frame(
    Sig = factor(
      ifelse(best_set[, "FDR"] < 0.05, "Yes", "No"),
      # TODO: temp trick to deal with the row-colouring problem
      # levels = c("Yes", "No")),
      levels = c("Yes")),
    row.names = rownames(best_set)),
  main = paste0("Cluster-group: ", chosen, " ", x, " - \n",
                y[1], "_top ", y[2], "_significance: ", y[3], " ; \n",
                z[1], "_top ", z[2], "_significance: ", z[3]),
  column_annotation_colors = list(
    # Sig = c("Yes" = "red", "No" = "lightgrey"),
    vs5 = vs5_colours,
    cluster = cluster_colours,
    stage = stage_colours,
    tissue = tissue_colours,
    donor = donor_colours,
    group = group_colours,
    plate_number = plate_number_colours),
  color = hcl.colors(101, "Blue-Red 3"),
  fontsize = 7)
```

DGE lists of these comparisons are available in [`output/marker_genes/whole_cell/uniquely_up/cluster_3_vs_4/`](..output/marker_genes/whole_cell/uniquely_up/cluster_3_vs_4/).

:::

::: {.panel}

#### `cluster_2_vs_cluster_3` {.panel-name}

```{r}
#########
# K vs L
#########

#################################################################################
# cluster 2 (i.e. S3-mix more Thymus) vs cluster 3 (i.e. mostly Thymus.S1 + .S2)

# checkpoint
cp <- sce
# exclude cells of uninterested cluster from cp
cp <- cp[, cp$cluster == "2" | cp$cluster == "3"]
colData(cp) <- droplevels(colData(cp))

# classify cluster-group for comparison
cp$vs6 <- factor(ifelse(cp$cluster == 2 , "K", "L"))

# set vs colours
vs6_colours <- setNames(
  palette.colors(nlevels(cp$vs6), "Set1"),
  levels(cp$vs6))
cp$colours$vs6_colours <- vs6_colours[cp$vs6]

# find unique DE ./. cluster-groups
vs6_uniquely_up <- findMarkers(
  cp,
  groups = cp$vs6,
  block = cp$block,
  pval.type = "all",
  direction = "up")

# export DGE lists
saveRDS(
  vs6_uniquely_up,
  here("data", "marker_genes", "whole_cell", "C094_Pellicci.uniquely_up.cluster_2_vs_3.rds"),
  compress = "xz")

dir.create(here("output", "marker_genes", "whole_cell", "uniquely_up", "cluster_2_vs_3"), recursive = TRUE)

vs_pair <- c("2", "3")

message("Writing 'uniquely_up (cluster_2_vs_3)' marker genes to file.")
for (n in names(vs6_uniquely_up)) {
  message(n)
  gzout <- gzfile(
    description = here(
      "output",
      "marker_genes",
      "whole_cell",
      "uniquely_up",
      "cluster_2_vs_3",
      paste0("cluster_",
             vs_pair[which(names(vs6_uniquely_up) %in% n)],
             "_vs_",
             vs_pair[-which(names(vs6_uniquely_up) %in% n)][1],
             ".uniquely_up.csv.gz")),
    open = "wb")
  write.table(
    x = vs6_uniquely_up[[n]] %>%
      as.data.frame() %>%
      tibble::rownames_to_column("gene_ID"),
    file = gzout,
    sep = ",",
    quote = FALSE,
    row.names = FALSE,
    col.names = TRUE)
  close(gzout)
}
```

```{r}
#####################################################################
# look at cluster-group K / cluster 2 (i.e. S3-mix more Thymus)
chosen <- "K"
K_uniquely_up <- vs6_uniquely_up[[chosen]]

# add description for the chosen cluster-group
x <- "(cluster 2; S3-mix more Thymus)"

# look only at protein coding gene (pcg)
# NOTE: not suggest to narrow down into pcg as it remove all significant candidates (FDR << 0.05) !
# K_uniquely_up_pcg <- K_uniquely_up[intersect(protein_coding_genK_set, rownames(K_uniquely_up)), ]

# get rid of noise (i.e. pseudo, ribo, mito, sex) that collaborator not interested in
K_uniquely_up_noiseR <- K_uniquely_up[setdiff(rownames(K_uniquely_up), c(pseudogene_set, mito_set, ribo_set, sex_set)), ]

# see if key marker, "CD4 and/or ""KLRB1/CD161"", contain in the DE list + if it is "significant (i.e FDR <0.05)
y <- c("CD4",
       which(rownames(K_uniquely_up_noiseR) %in% "CD4"),
       K_uniquely_up_noiseR[which(rownames(K_uniquely_up_noiseR) %in% "CD4"), ]$FDR < 0.05)
z <- c("KLRB1/CD161",
       which(rownames(K_uniquely_up_noiseR) %in% "KLRB1"),
       K_uniquely_up_noiseR[which(rownames(K_uniquely_up_noiseR) %in% "KLRB1"), ]$FDR < 0.05)

# top25 only + gene-of-interest
best_set <- K_uniquely_up_noiseR[1:25, ]
```

```{r heat-uniquely-up-logExp-cluster-2-vs-3, fig.cap = "Heatmap of log-expression values in each sample for the top uniquely upregulated marker genes. Each column is a sample, each row a gene. Colours are capped at -3 and 3 to preserve dynamic range. Ranking of CD4 and CD161/KLRB1 from top of the DGE list sorted in ascending order of FDR and their statistical significance (TRUE = FDR < 0.05) are provided in the title.", fig.asp = 1.2, layout="l-body-outset"}
# heatmap
plotHeatmap(
  cp,
  features = rownames(best_set),
  columns = order(
    cp$vs6,
    cp$cluster,
    cp$stage,
    cp$tissue,
    cp$donor,
    cp$group,
    cp$plate_number),
  colour_columns_by = c(
    "vs6",
    "cluster",
    "stage",
    "tissue",
    "donor",
    "group",
    "plate_number"),
  cluster_cols = FALSE,
  center = TRUE,
  symmetric = TRUE,
  zlim = c(-3, 3),
  show_colnames = FALSE,
  annotation_row = data.frame(
    Sig = factor(
      ifelse(best_set[, "FDR"] < 0.05, "Yes", "No"),
      # TODO: temp trick to deal with the row-colouring problem
      # levels = c("Yes", "No")),
      levels = c("Yes")),
    row.names = rownames(best_set)),
  main = paste0("Cluster-group: ", chosen, " ", x, " - \n",
                y[1], "_top ", y[2], "_significance: ", y[3], " ; \n",
                z[1], "_top ", z[2], "_significance: ", z[3]),
  column_annotation_colors = list(
    # Sig = c("Yes" = "red", "No" = "lightgrey"),
    vs6 = vs6_colours,
    cluster = cluster_colours,
    stage = stage_colours,
    tissue = tissue_colours,
    donor = donor_colours,
    group = group_colours,
    plate_number = plate_number_colours),
  color = hcl.colors(101, "Blue-Red 3"),
  fontsize = 7)
```

```{r}
##############################################################
# look at cluster-group L / cluster 3  (i.e. mostly Thymus.S1 + .S2)
chosen <- "L"
L_uniquely_up <- vs6_uniquely_up[[chosen]]

# add description for the chosen cluster-group
x <- "(cluster 3; mostly Thymus.S3)"

# look only at protein coding gene (pcg)
# NOTE: not suggest to narrow down into pcg as it remove all significant candidates (FDR << 0.05) !
# L_uniquely_up_pcg <- L_uniquely_up[intersect(protein_coding_genK_set, rownames(L_uniquely_up)), ]

# get rid of noise (i.e. pseudo, ribo, mito, sex) that collaborator not interested in
L_uniquely_up_noiseR <- L_uniquely_up[setdiff(rownames(L_uniquely_up), c(pseudogene_set, mito_set, ribo_set, sex_set)), ]

# see if key marker, "CD4 and/or ""KLRB1/CD161"", contain in the DE list + if it is "significant (i.e FDR <0.05)
y <- c("CD4",
       which(rownames(L_uniquely_up_noiseR) %in% "CD4"),
       L_uniquely_up_noiseR[which(rownames(L_uniquely_up_noiseR) %in% "CD4"), ]$FDR < 0.05)
z <- c("KLRB1/CD161",
       which(rownames(L_uniquely_up_noiseR) %in% "KLRB1"),
       L_uniquely_up_noiseR[which(rownames(L_uniquely_up_noiseR) %in% "KLRB1"), ]$FDR < 0.05)

# top25 only + gene-of-interest
best_set <- L_uniquely_up_noiseR[1:25, ]
```

```{r heat-uniquely-up-logExp-cluster-3-vs-2, fig.cap = "Heatmap of log-expression values in each sample for the top uniquely upregulated marker genes. Each column is a sample, each row a gene. Colours are capped at -3 and 3 to preserve dynamic range. Ranking of CD4 and CD161/KLRB1 from top of the DGE list sorted in ascending order of FDR and their statistical significance (TRUE = FDR < 0.05) are provided in the title.", fig.asp = 1.2, layout="l-body-outset"}
# heatmap
plotHeatmap(
  cp,
  features = rownames(best_set),
  columns = order(
    cp$vs6,
    cp$cluster,
    cp$stage,
    cp$tissue,
    cp$donor,
    cp$group,
    cp$plate_number),
  colour_columns_by = c(
    "vs6",
    "cluster",
    "stage",
    "tissue",
    "donor",
    "group",
    "plate_number"),
  cluster_cols = FALSE,
  center = TRUE,
  symmetric = TRUE,
  zlim = c(-3, 3),
  show_colnames = FALSE,
  annotation_row = data.frame(
    Sig = factor(
      ifelse(best_set[, "FDR"] < 0.05, "Yes", "No"),
      # TODO: temp trick to deal with the row-colouring problem
      # levels = c("Yes", "No")),
      levels = c("Yes")),
    row.names = rownames(best_set)),
  main = paste0("Cluster-group: ", chosen, " ", x, " - \n",
                y[1], "_top ", y[2], "_significance: ", y[3], " ; \n",
                z[1], "_top ", z[2], "_significance: ", z[3]),
  column_annotation_colors = list(
    # Sig = c("Yes" = "red", "No" = "lightgrey"),
    vs6 = vs6_colours,
    cluster = cluster_colours,
    stage = stage_colours,
    tissue = tissue_colours,
    donor = donor_colours,
    group = group_colours,
    plate_number = plate_number_colours),
  color = hcl.colors(101, "Blue-Red 3"),
  fontsize = 7)
```

DGE lists of these comparisons are available in [`output/marker_genes/whole_cell/uniquely_up/cluster_2_vs_3/`](..output/marker_genes/whole_cell/uniquely_up/cluster_2_vs_3/).

:::

::: {.panel}

#### `cluster_2_vs_cluster_4` {.panel-name}

```{r}
#########
# M vs N
#########

#################################################################################
# cluster 2 (i.e. S3-mix more Thymus) vs cluster 4 (i.e. mostly Thymus.S3)

# checkpoint
cp <- sce
# exclude cells of uninterested cluster from cp
cp <- cp[, cp$cluster == "2" | cp$cluster == "4"]
colData(cp) <- droplevels(colData(cp))

# classify cluster-group for comparison
cp$vs7 <- factor(ifelse(cp$cluster == 2 , "M", "N"))

# set vs colours
vs7_colours <- setNames(
  palette.colors(nlevels(cp$vs7), "Set1"),
  levels(cp$vs7))
cp$colours$vs7_colours <- vs7_colours[cp$vs7]

# find unique DE ./. cluster-groups
vs7_uniquely_up <- findMarkers(
  cp,
  groups = cp$vs7,
  block = cp$block,
  pval.type = "all",
  direction = "up")

# export DGE lists
saveRDS(
  vs7_uniquely_up,
  here("data", "marker_genes", "whole_cell", "C094_Pellicci.uniquely_up.cluster_2_vs_4.rds"),
  compress = "xz")

dir.create(here("output", "marker_genes", "whole_cell", "uniquely_up", "cluster_2_vs_4"), recursive = TRUE)

vs_pair <- c("2", "4")

message("Writing 'uniquely_up (cluster_2_vs_4)' marker genes to file.")
for (n in names(vs7_uniquely_up)) {
  message(n)
  gzout <- gzfile(
    description = here(
      "output",
      "marker_genes",
      "whole_cell",
      "uniquely_up",
      "cluster_2_vs_4",
      paste0("cluster_",
             vs_pair[which(names(vs7_uniquely_up) %in% n)],
             "_vs_",
             vs_pair[-which(names(vs7_uniquely_up) %in% n)][1],
             ".uniquely_up.csv.gz")),
    open = "wb")
  write.table(
    x = vs7_uniquely_up[[n]] %>%
      as.data.frame() %>%
      tibble::rownames_to_column("gene_ID"),
    file = gzout,
    sep = ",",
    quote = FALSE,
    row.names = FALSE,
    col.names = TRUE)
  close(gzout)
}
```

```{r}
#####################################################################
# look at cluster-group M / cluster 2 (i.e. S3-mix more Thymus)
chosen <- "M"
M_uniquely_up <- vs7_uniquely_up[[chosen]]

# add description for the chosen cluster-group
x <- "(cluster 2; S3-mix more Thymus)"

# look only at protein coding gene (pcg)
# NOTE: not suggest to narrow down into pcg as it remove all significant candidates (FDR << 0.05) !
# M_uniquely_up_pcg <- M_uniquely_up[intersect(protein_coding_genM_set, rownames(M_uniquely_up)), ]

# get rid of noise (i.e. pseudo, ribo, mito, sex) that collaborator not interested in
M_uniquely_up_noiseR <- M_uniquely_up[setdiff(rownames(M_uniquely_up), c(pseudogene_set, mito_set, ribo_set, sex_set)), ]

# see if key marker, "CD4 and/or ""KLRB1/CD161"", contain in the DE list + if it is "significant (i.e FDR <0.05)
y <- c("CD4",
       which(rownames(M_uniquely_up_noiseR) %in% "CD4"),
       M_uniquely_up_noiseR[which(rownames(M_uniquely_up_noiseR) %in% "CD4"), ]$FDR < 0.05)
z <- c("KLRB1/CD161",
       which(rownames(M_uniquely_up_noiseR) %in% "KLRB1"),
       M_uniquely_up_noiseR[which(rownames(M_uniquely_up_noiseR) %in% "KLRB1"), ]$FDR < 0.05)

# top25 only + gene-of-interest
best_set <- M_uniquely_up_noiseR[1:25, ]
```

```{r heat-uniquely-up-logExp-cluster-2-vs-4, fig.cap = "Heatmap of log-expression values in each sample for the top uniquely upregulated marker genes. Each column is a sample, each row a gene. Colours are capped at -3 and 3 to preserve dynamic range. Ranking of CD4 and CD161/KLRB1 from top of the DGE list sorted in ascending order of FDR and their statistical significance (TRUE = FDR < 0.05) are provided in the title.", fig.asp = 1.2, layout="l-body-outset"}
# heatmap
plotHeatmap(
  cp,
  features = rownames(best_set),
  columns = order(
    cp$vs7,
    cp$cluster,
    cp$stage,
    cp$tissue,
    cp$donor,
    cp$group,
    cp$plate_number),
  colour_columns_by = c(
    "vs7",
    "cluster",
    "stage",
    "tissue",
    "donor",
    "group",
    "plate_number"),
  cluster_cols = FALSE,
  center = TRUE,
  symmetric = TRUE,
  zlim = c(-3, 3),
  show_colnames = FALSE,
  annotation_row = data.frame(
    Sig = factor(
      ifelse(best_set[, "FDR"] < 0.05, "Yes", "No"),
      # TODO: temp trick to deal with the row-colouring problem
      # levels = c("Yes", "No")),
      levels = c("Yes")),
    row.names = rownames(best_set)),
  main = paste0("Cluster-group: ", chosen, " ", x, " - \n",
                y[1], "_top ", y[2], "_significance: ", y[3], " ; \n",
                z[1], "_top ", z[2], "_significance: ", z[3]),
  column_annotation_colors = list(
    # Sig = c("Yes" = "red", "No" = "lightgrey"),
    vs7 = vs7_colours,
    cluster = cluster_colours,
    stage = stage_colours,
    tissue = tissue_colours,
    donor = donor_colours,
    group = group_colours,
    plate_number = plate_number_colours),
  color = hcl.colors(101, "Blue-Red 3"),
  fontsize = 7)
```

```{r}
##############################################################
# look at cluster-group N / cluster 4  (i.e. mostly Thymus.S3)
chosen <- "N"
N_uniquely_up <- vs7_uniquely_up[[chosen]]

# add description for the chosen cluster-group
x <- "(cluster 4; mostly Thymus.S3)"

# look only at protein coding gene (pcg)
# NOTE: not suggest to narrow down into pcg as it remove all significant candidates (FDR << 0.05) !
# N_uniquely_up_pcg <- N_uniquely_up[intersect(protein_coding_genM_set, rownames(N_uniquely_up)), ]

# get rid of noise (i.e. pseudo, ribo, mito, sex) that collaborator not interested in
N_uniquely_up_noiseR <- N_uniquely_up[setdiff(rownames(N_uniquely_up), c(pseudogene_set, mito_set, ribo_set, sex_set)), ]

# see if key marker, "CD4 and/or ""KLRB1/CD161"", contain in the DE list + if it is "significant (i.e FDR <0.05)
y <- c("CD4",
       which(rownames(N_uniquely_up_noiseR) %in% "CD4"),
       N_uniquely_up_noiseR[which(rownames(N_uniquely_up_noiseR) %in% "CD4"), ]$FDR < 0.05)
z <- c("KLRB1/CD161",
       which(rownames(N_uniquely_up_noiseR) %in% "KLRB1"),
       N_uniquely_up_noiseR[which(rownames(N_uniquely_up_noiseR) %in% "KLRB1"), ]$FDR < 0.05)

# top25 only + gene-of-interest
best_set <- N_uniquely_up_noiseR[1:25, ]
```

```{r heat-uniquely-up-logExp-cluster-4-vs-2, fig.cap = "Heatmap of log-expression values in each sample for the top uniquely upregulated marker genes. Each column is a sample, each row a gene. Colours are capped at -3 and 3 to preserve dynamic range. Ranking of CD4 and CD161/KLRB1 from top of the DGE list sorted in ascending order of FDR and their statistical significance (TRUE = FDR < 0.05) are provided in the title.", fig.asp = 1.2, layout="l-body-outset"}
# heatmap
plotHeatmap(
  cp,
  features = rownames(best_set),
  columns = order(
    cp$vs7,
    cp$cluster,
    cp$stage,
    cp$tissue,
    cp$donor,
    cp$group,
    cp$plate_number),
  colour_columns_by = c(
    "vs7",
    "cluster",
    "stage",
    "tissue",
    "donor",
    "group",
    "plate_number"),
  cluster_cols = FALSE,
  center = TRUE,
  symmetric = TRUE,
  zlim = c(-3, 3),
  show_colnames = FALSE,
  annotation_row = data.frame(
    Sig = factor(
      ifelse(best_set[, "FDR"] < 0.05, "Yes", "No"),
      # TODO: temp trick to deal with the row-colouring problem
      # levels = c("Yes", "No")),
      levels = c("Yes")),
    row.names = rownames(best_set)),
  main = paste0("Cluster-group: ", chosen, " ", x, " - \n",
                y[1], "_top ", y[2], "_significance: ", y[3], " ; \n",
                z[1], "_top ", z[2], "_significance: ", z[3]),
  column_annotation_colors = list(
    # Sig = c("Yes" = "red", "No" = "lightgrey"),
    vs7 = vs7_colours,
    cluster = cluster_colours,
    stage = stage_colours,
    tissue = tissue_colours,
    donor = donor_colours,
    group = group_colours,
    plate_number = plate_number_colours),
  color = hcl.colors(101, "Blue-Red 3"),
  fontsize = 7)
```

DGE lists of these comparisons are available in [`output/marker_genes/whole_cell/uniquely_up/cluster_2_vs_4/`](..output/marker_genes/whole_cell/uniquely_up/cluster_2_vs_4/).

:::

::: {.panel}

#### `cluster_1_vs_cluster_3` {.panel-name}

```{r}
#########
# O vs P
#########

#################################################################################
# cluster 1 (i.e. S3-mix more Blood) vs cluster 3 (i.e. mostly Thymus.S1 and .S2)

# checkpoint
cp <- sce
# exclude cells of uninterested cluster from cp
cp <- cp[, cp$cluster == "1" | cp$cluster == "3"]
colData(cp) <- droplevels(colData(cp))

# classify cluster-group for comparison
cp$vs8 <- factor(ifelse(cp$cluster == 1 , "O", "P"))

# set vs colours
vs8_colours <- setNames(
  palette.colors(nlevels(cp$vs8), "Set1"),
  levels(cp$vs8))
cp$colours$vs8_colours <- vs8_colours[cp$vs8]

# find unique DE ./. cluster-groups
vs8_uniquely_up <- findMarkers(
  cp,
  groups = cp$vs8,
  block = cp$block,
  pval.type = "all",
  direction = "up")

# export DGE lists
saveRDS(
  vs8_uniquely_up,
  here("data", "marker_genes", "whole_cell", "C094_Pellicci.uniquely_up.cluster_1_vs_3.rds"),
  compress = "xz")

dir.create(here("output", "marker_genes", "whole_cell", "uniquely_up", "cluster_1_vs_3"), recursive = TRUE)

vs_pair <- c("1", "3")

message("Writing 'uniquely_up (cluster_1_vs_3)' marker genes to file.")
for (n in names(vs8_uniquely_up)) {
  message(n)
  gzout <- gzfile(
    description = here(
      "output",
      "marker_genes",
      "whole_cell",
      "uniquely_up",
      "cluster_1_vs_3",
      paste0("cluster_",
             vs_pair[which(names(vs8_uniquely_up) %in% n)],
             "_vs_",
             vs_pair[-which(names(vs8_uniquely_up) %in% n)][1],
             ".uniquely_up.csv.gz")),
    open = "wb")
  write.table(
    x = vs8_uniquely_up[[n]] %>%
      as.data.frame() %>%
      tibble::rownames_to_column("gene_ID"),
    file = gzout,
    sep = ",",
    quote = FALSE,
    row.names = FALSE,
    col.names = TRUE)
  close(gzout)
}
```

```{r}
#####################################################################
# look at cluster-group O / cluster 1 (i.e. S3-mix more Blood)
chosen <- "O"
O_uniquely_up <- vs8_uniquely_up[[chosen]]

# add description for the chosen cluster-group
x <- "(cluster 1; S3-mix more Blood)"

# look only at protein coding gene (pcg)
# NOTE: not suggest to narrow down into pcg as it remove all significant candidates (FDR << 0.05) !
# O_uniquely_up_pcg <- O_uniquely_up[intersect(protein_coding_genO_set, rownames(O_uniquely_up)), ]

# get rid of noise (i.e. pseudo, ribo, mito, sex) that collaborator not interested in
O_uniquely_up_noiseR <- O_uniquely_up[setdiff(rownames(O_uniquely_up), c(pseudogene_set, mito_set, ribo_set, sex_set)), ]

# see if key marker, "CD4 and/or ""KLRB1/CD161"", contain in the DE list + if it is "significant (i.e FDR <0.05)
y <- c("CD4",
       which(rownames(O_uniquely_up_noiseR) %in% "CD4"),
       O_uniquely_up_noiseR[which(rownames(O_uniquely_up_noiseR) %in% "CD4"), ]$FDR < 0.05)
z <- c("KLRB1/CD161",
       which(rownames(O_uniquely_up_noiseR) %in% "KLRB1"),
       O_uniquely_up_noiseR[which(rownames(O_uniquely_up_noiseR) %in% "KLRB1"), ]$FDR < 0.05)

# top25 only + gene-of-interest
best_set <- O_uniquely_up_noiseR[1:25, ]
```

```{r heat-uniquely-up-logExp-cluster-1-vs-3, fig.cap = "Heatmap of log-expression values in each sample for the top uniquely upregulated marker genes. Each column is a sample, each row a gene. Colours are capped at -3 and 3 to preserve dynamic range. Ranking of CD4 and CD161/KLRB1 from top of the DGE list sorted in ascending order of FDR and their statistical significance (TRUE = FDR < 0.05) are provided in the title.", fig.asp = 1.2, layout="l-body-outset"}
# heatmap
plotHeatmap(
  cp,
  features = rownames(best_set),
  columns = order(
    cp$vs8,
    cp$cluster,
    cp$stage,
    cp$tissue,
    cp$donor,
    cp$group,
    cp$plate_number),
  colour_columns_by = c(
    "vs8",
    "cluster",
    "stage",
    "tissue",
    "donor",
    "group",
    "plate_number"),
  cluster_cols = FALSE,
  center = TRUE,
  symmetric = TRUE,
  zlim = c(-3, 3),
  show_colnames = FALSE,
  annotation_row = data.frame(
    Sig = factor(
      ifelse(best_set[, "FDR"] < 0.05, "Yes", "No"),
      # TODO: temp trick to deal with the row-colouring problem
      # levels = c("Yes", "No")),
      levels = c("Yes")),
    row.names = rownames(best_set)),
  main = paste0("Cluster-group: ", chosen, " ", x, " - \n",
                y[1], "_top ", y[2], "_significance: ", y[3], " ; \n",
                z[1], "_top ", z[2], "_significance: ", z[3]),
  column_annotation_colors = list(
    # Sig = c("Yes" = "red", "No" = "lightgrey"),
    vs8 = vs8_colours,
    cluster = cluster_colours,
    stage = stage_colours,
    tissue = tissue_colours,
    donor = donor_colours,
    group = group_colours,
    plate_number = plate_number_colours),
  color = hcl.colors(101, "Blue-Red 3"),
  fontsize = 7)
```

```{r}
##############################################################
# look at cluster-group P / cluster 3  (i.e. mostly Thymus.S1 and .S2)
chosen <- "P"
P_uniquely_up <- vs8_uniquely_up[[chosen]]

# add description for the chosen cluster-group
x <- "(cluster 3; mostly Thymus.S1 and .S2)"

# look only at protein coding gene (pcg)
# NOTE: not suggest to narrow down into pcg as it remove all significant candidates (FDR << 0.05) !
# P_uniquely_up_pcg <- P_uniquely_up[intersect(protein_coding_genO_set, rownames(P_uniquely_up)), ]

# get rid of noise (i.e. pseudo, ribo, mito, sex) that collaborator not interested in
P_uniquely_up_noiseR <- P_uniquely_up[setdiff(rownames(P_uniquely_up), c(pseudogene_set, mito_set, ribo_set, sex_set)), ]

# see if key marker, "CD4 and/or ""KLRB1/CD161"", contain in the DE list + if it is "significant (i.e FDR <0.05)
y <- c("CD4",
       which(rownames(P_uniquely_up_noiseR) %in% "CD4"),
       P_uniquely_up_noiseR[which(rownames(P_uniquely_up_noiseR) %in% "CD4"), ]$FDR < 0.05)
z <- c("KLRB1/CD161",
       which(rownames(P_uniquely_up_noiseR) %in% "KLRB1"),
       P_uniquely_up_noiseR[which(rownames(P_uniquely_up_noiseR) %in% "KLRB1"), ]$FDR < 0.05)

# top25 only + gene-of-interest
best_set <- P_uniquely_up_noiseR[1:25, ]
```

```{r heat-uniquely-up-logExp-cluster-3-vs-1, fig.cap = "Heatmap of log-expression values in each sample for the top uniquely upregulated marker genes. Each column is a sample, each row a gene. Colours are capped at -3 and 3 to preserve dynamic range. Ranking of CD4 and CD161/KLRB1 from top of the DGE list sorted in ascending order of FDR and their statistical significance (TRUE = FDR < 0.05) are provided in the title.", fig.asp = 1.2, layout="l-body-outset"}
# heatmap
plotHeatmap(
  cp,
  features = rownames(best_set),
  columns = order(
    cp$vs8,
    cp$cluster,
    cp$stage,
    cp$tissue,
    cp$donor,
    cp$group,
    cp$plate_number),
  colour_columns_by = c(
    "vs8",
    "cluster",
    "stage",
    "tissue",
    "donor",
    "group",
    "plate_number"),
  cluster_cols = FALSE,
  center = TRUE,
  symmetric = TRUE,
  zlim = c(-3, 3),
  show_colnames = FALSE,
  annotation_row = data.frame(
    Sig = factor(
      ifelse(best_set[, "FDR"] < 0.05, "Yes", "No"),
      # TODO: temp trick to deal with the row-colouring problem
      # levels = c("Yes", "No")),
      levels = c("Yes")),
    row.names = rownames(best_set)),
  main = paste0("Cluster-group: ", chosen, " ", x, " - \n",
                y[1], "_top ", y[2], "_significance: ", y[3], " ; \n",
                z[1], "_top ", z[2], "_significance: ", z[3]),
  column_annotation_colors = list(
    # Sig = c("Yes" = "red", "No" = "lightgrey"),
    vs8 = vs8_colours,
    cluster = cluster_colours,
    stage = stage_colours,
    tissue = tissue_colours,
    donor = donor_colours,
    group = group_colours,
    plate_number = plate_number_colours),
  color = hcl.colors(101, "Blue-Red 3"),
  fontsize = 7)
```

DGE lists of these comparisons are available in [`output/marker_genes/whole_cell/uniquely_up/cluster_1_vs_3/`](..output/marker_genes/whole_cell/uniquely_up/cluster_1_vs_3/).

:::

::: {.panel}

#### `cluster_1_vs_cluster_4` {.panel-name}

```{r}
#########
# Q vs R
#########

#################################################################################
# cluster 1 (i.e. S3-mix more Blood) vs cluster 4 (i.e. mostly Thymus.S3)

# checkpoint
cp <- sce
# exclude cells of uninterested cluster from cp
cp <- cp[, cp$cluster == "1" | cp$cluster == "4"]
colData(cp) <- droplevels(colData(cp))

# classify cluster-group for comparison
cp$vs9 <- factor(ifelse(cp$cluster == 1 , "Q", "R"))

# set vs colours
vs9_colours <- setNames(
  palette.colors(nlevels(cp$vs9), "Set1"),
  levels(cp$vs9))
cp$colours$vs9_colours <- vs9_colours[cp$vs9]

# find unique DE ./. cluster-groups
vs9_uniquely_up <- findMarkers(
  cp,
  groups = cp$vs9,
  block = cp$block,
  pval.type = "all",
  direction = "up")

# export DGE lists
saveRDS(
  vs9_uniquely_up,
  here("data", "marker_genes", "whole_cell", "C094_Pellicci.uniquely_up.cluster_1_vs_4.rds"),
  compress = "xz")

dir.create(here("output", "marker_genes", "whole_cell", "uniquely_up", "cluster_1_vs_4"), recursive = TRUE)

vs_pair <- c("1", "4")

message("Writing 'uniquely_up (cluster_1_vs_4)' marker genes to file.")
for (n in names(vs9_uniquely_up)) {
  message(n)
  gzout <- gzfile(
    description = here(
      "output",
      "marker_genes",
      "whole_cell",
      "uniquely_up",
      "cluster_1_vs_4",
      paste0("cluster_",
             vs_pair[which(names(vs9_uniquely_up) %in% n)],
             "_vs_",
             vs_pair[-which(names(vs9_uniquely_up) %in% n)][1],
             ".uniquely_up.csv.gz")),
    open = "wb")
  write.table(
    x = vs9_uniquely_up[[n]] %>%
      as.data.frame() %>%
      tibble::rownames_to_column("gene_ID"),
    file = gzout,
    sep = ",",
    quote = FALSE,
    row.names = FALSE,
    col.names = TRUE)
  close(gzout)
}
```

```{r}
#####################################################################
# look at cluster-group Q / cluster 1 (i.e. S3-mix more Blood)
chosen <- "Q"
Q_uniquely_up <- vs9_uniquely_up[[chosen]]

# add description for the chosen cluster-group
x <- "(cluster 1; S3-mix more Blood)"

# look only at protein coding gene (pcg)
# NOTE: not suggest to narrow down into pcg as it remove all significant candidates (FDR << 0.05) !
# Q_uniquely_up_pcg <- Q_uniquely_up[intersect(protein_coding_genQ_set, rownames(Q_uniquely_up)), ]

# get rid of noise (i.e. pseudo, ribo, mito, sex) that collaborator not interested in
Q_uniquely_up_noiseR <- Q_uniquely_up[setdiff(rownames(Q_uniquely_up), c(pseudogene_set, mito_set, ribo_set, sex_set)), ]

# see if key marker, "CD4 and/or ""KLRB1/CD161"", contain in the DE list + if it is "significant (i.e FDR <0.05)
y <- c("CD4",
       which(rownames(Q_uniquely_up_noiseR) %in% "CD4"),
       Q_uniquely_up_noiseR[which(rownames(Q_uniquely_up_noiseR) %in% "CD4"), ]$FDR < 0.05)
z <- c("KLRB1/CD161",
       which(rownames(Q_uniquely_up_noiseR) %in% "KLRB1"),
       Q_uniquely_up_noiseR[which(rownames(Q_uniquely_up_noiseR) %in% "KLRB1"), ]$FDR < 0.05)

# top25 only + gene-of-interest
best_set <- Q_uniquely_up_noiseR[1:25, ]
```

```{r heat-uniquely-up-logExp-cluster-1-vs-4, fig.cap = "Heatmap of log-expression values in each sample for the top uniquely upregulated marker genes. Each column is a sample, each row a gene. Colours are capped at -3 and 3 to preserve dynamic range. Ranking of CD4 and CD161/KLRB1 from top of the DGE list sorted in ascending order of FDR and their statistical significance (TRUE = FDR < 0.05) are provided in the title.", fig.asp = 1.2, layout="l-body-outset"}
# heatmap
plotHeatmap(
  cp,
  features = rownames(best_set),
  columns = order(
    cp$vs9,
    cp$cluster,
    cp$stage,
    cp$tissue,
    cp$donor,
    cp$group,
    cp$plate_number),
  colour_columns_by = c(
    "vs9",
    "cluster",
    "stage",
    "tissue",
    "donor",
    "group",
    "plate_number"),
  cluster_cols = FALSE,
  center = TRUE,
  symmetric = TRUE,
  zlim = c(-3, 3),
  show_colnames = FALSE,
  annotation_row = data.frame(
    Sig = factor(
      ifelse(best_set[, "FDR"] < 0.05, "Yes", "No"),
      # TODO: temp trick to deal with the row-colouring problem
      # levels = c("Yes", "No")),
      levels = c("Yes")),
    row.names = rownames(best_set)),
  main = paste0("Cluster-group: ", chosen, " ", x, " - \n",
                y[1], "_top ", y[2], "_significance: ", y[3], " ; \n",
                z[1], "_top ", z[2], "_significance: ", z[3]),
  column_annotation_colors = list(
    # Sig = c("Yes" = "red", "No" = "lightgrey"),
    vs9 = vs9_colours,
    cluster = cluster_colours,
    stage = stage_colours,
    tissue = tissue_colours,
    donor = donor_colours,
    group = group_colours,
    plate_number = plate_number_colours),
  color = hcl.colors(101, "Blue-Red 3"),
  fontsize = 7)
```

```{r}
##############################################################
# look at cluster-group R / cluster 4  (i.e. mostly Thymus.S3)
chosen <- "R"
R_uniquely_up <- vs9_uniquely_up[[chosen]]

# add description for the chosen cluster-group
x <- "(cluster 4; mostly Thymus.S3)"

# look only at protein coding gene (pcg)
# NOTE: not suggest to narrow down into pcg as it remove all significant candidates (FDR << 0.05) !
# R_uniquely_up_pcg <- R_uniquely_up[intersect(protein_coding_genQ_set, rownames(R_uniquely_up)), ]

# get rid of noise (i.e. pseudo, ribo, mito, sex) that collaborator not interested in
R_uniquely_up_noiseR <- R_uniquely_up[setdiff(rownames(R_uniquely_up), c(pseudogene_set, mito_set, ribo_set, sex_set)), ]

# see if key marker, "CD4 and/or ""KLRB1/CD161"", contain in the DE list + if it is "significant (i.e FDR <0.05)
y <- c("CD4",
       which(rownames(R_uniquely_up_noiseR) %in% "CD4"),
       R_uniquely_up_noiseR[which(rownames(R_uniquely_up_noiseR) %in% "CD4"), ]$FDR < 0.05)
z <- c("KLRB1/CD161",
       which(rownames(R_uniquely_up_noiseR) %in% "KLRB1"),
       R_uniquely_up_noiseR[which(rownames(R_uniquely_up_noiseR) %in% "KLRB1"), ]$FDR < 0.05)

# top25 only + gene-of-interest
best_set <- R_uniquely_up_noiseR[1:25, ]
```

```{r heat-uniquely-up-logExp-cluster-4-vs-1, fig.cap = "Heatmap of log-expression values in each sample for the top uniquely upregulated marker genes. Each column is a sample, each row a gene. Colours are capped at -3 and 3 to preserve dynamic range. Ranking of CD4 and CD161/KLRB1 from top of the DGE list sorted in ascending order of FDR and their statistical significance (TRUE = FDR < 0.05) are provided in the title.", fig.asp = 1.2, layout="l-body-outset"}
# heatmap
plotHeatmap(
  cp,
  features = rownames(best_set),
  columns = order(
    cp$vs9,
    cp$cluster,
    cp$stage,
    cp$tissue,
    cp$donor,
    cp$group,
    cp$plate_number),
  colour_columns_by = c(
    "vs9",
    "cluster",
    "stage",
    "tissue",
    "donor",
    "group",
    "plate_number"),
  cluster_cols = FALSE,
  center = TRUE,
  symmetric = TRUE,
  zlim = c(-3, 3),
  show_colnames = FALSE,
  annotation_row = data.frame(
    Sig = factor(
      ifelse(best_set[, "FDR"] < 0.05, "Yes", "No"),
      # TODO: temp trick to deal with the row-colouring problem
      # levels = c("Yes", "No")),
      levels = c("Yes")),
    row.names = rownames(best_set)),
  main = paste0("Cluster-group: ", chosen, " ", x, " - \n",
                y[1], "_top ", y[2], "_significance: ", y[3], " ; \n",
                z[1], "_top ", z[2], "_significance: ", z[3]),
  column_annotation_colors = list(
    # Sig = c("Yes" = "red", "No" = "lightgrey"),
    vs9 = vs9_colours,
    cluster = cluster_colours,
    stage = stage_colours,
    tissue = tissue_colours,
    donor = donor_colours,
    group = group_colours,
    plate_number = plate_number_colours),
  color = hcl.colors(101, "Blue-Red 3"),
  fontsize = 7)
```

DGE lists of these comparisons are available in [`output/marker_genes/whole_cell/uniquely_up/cluster_1_vs_4/`](..output/marker_genes/whole_cell/uniquely_up/cluster_1_vs_4/).

:::

::::::

## Selected pairwise comparisons - integrated

To have a comprehensive overview, the above analyses were summarized in the following heatmap per cluster:

```{r, echo = FALSE}
# NOTE: The following is a workaround to the lack of support for tabsets in 
#       distill (see https://github.com/rstudio/distill/issues/11 and 
#       https://github.com/rstudio/distill/issues/11#issuecomment-692142414 in 
#       particular).
xaringanExtra::use_panelset()
```

::::: {.panelset}

::: {.panel}

#### `Cluster_3_integrated` {.panel-name}

```{r}
############
# cluster 3
############

chosen <- "3"

# add description for the chosen cluster-group
x <- "(mostly Thymus.S1 and Thymus.S2)"

# retain only significant markers (FDR<0.05) + keep only required output columns
B_uniquely_up_noiseR_sig <- B_uniquely_up_noiseR[B_uniquely_up_noiseR$FDR<0.05,][ ,c(1:3)]
D_uniquely_up_noiseR_sig <- D_uniquely_up_noiseR[D_uniquely_up_noiseR$FDR<0.05,][ ,c(1:3)]
I_uniquely_up_noiseR_sig <- I_uniquely_up_noiseR[I_uniquely_up_noiseR$FDR<0.05,][ ,c(1:3)]
L_uniquely_up_noiseR_sig <- L_uniquely_up_noiseR[L_uniquely_up_noiseR$FDR<0.05,][ ,c(1:3)]
P_uniquely_up_noiseR_sig <- P_uniquely_up_noiseR[P_uniquely_up_noiseR$FDR<0.05,][ ,c(1:3)]
cluster3_uniquely_up_noiseR_sig <- cluster3_uniquely_up_noiseR[cluster3_uniquely_up_noiseR$FDR<0.05,][ ,c(1:3)]

# add top column
B_uniquely_up_noiseR_sig$top <- 1:nrow(B_uniquely_up_noiseR_sig)
D_uniquely_up_noiseR_sig$top <- 1:nrow(D_uniquely_up_noiseR_sig)
I_uniquely_up_noiseR_sig$top <- 1:nrow(I_uniquely_up_noiseR_sig)
L_uniquely_up_noiseR_sig$top <- 1:nrow(L_uniquely_up_noiseR_sig)
P_uniquely_up_noiseR_sig$top <- 1:nrow(P_uniquely_up_noiseR_sig)
cluster3_uniquely_up_noiseR_sig$top <- 1:nrow(cluster3_uniquely_up_noiseR_sig)

# unify S4 objects, sort by top (ascending) then FDR (ascending), keep only first unique entry for each marker
BDILP3_uniquely_up_noiseR_sig <- rbind2(B_uniquely_up_noiseR_sig,
                                        D_uniquely_up_noiseR_sig,
                                        I_uniquely_up_noiseR_sig,
                                        L_uniquely_up_noiseR_sig,
                                        P_uniquely_up_noiseR_sig,
                                        cluster3_uniquely_up_noiseR_sig)

BDILP3_uniquely_up_noiseR_sig_sort <- BDILP3_uniquely_up_noiseR_sig[with(BDILP3_uniquely_up_noiseR_sig, order(top, FDR)), ]
BDILP3_uniquely_up_noiseR_sig_sort_uniq <- BDILP3_uniquely_up_noiseR_sig_sort[unique(rownames(BDILP3_uniquely_up_noiseR_sig_sort)), ]

# # de-select unannotated/ not well-characterised genes
# deselected <- c("NPIPB13", "NPIPB3", "NPIPB11", "NPIPB5", "NPIPB4", "EEF1A1", "ACTG1", "ACTB", "IFITM1")
deselected <- c()
BDILP3_uniquely_up_noiseR_sig_sort_uniq_selected <- BDILP3_uniquely_up_noiseR_sig_sort_uniq[!(rownames(BDILP3_uniquely_up_noiseR_sig_sort_uniq) %in% deselected), ]

# export DGE lists
saveRDS(
  BDILP3_uniquely_up_noiseR_sig_sort_uniq_selected,
  here("data", "marker_genes", "whole_cell", "C094_Pellicci.uniquely_up.cluster_3_integrated.rds"),
  compress = "xz")

dir.create(here("output", "marker_genes", "whole_cell", "uniquely_up", "integrated"), recursive = TRUE)

message("Writing 'uniquely_up (cluster_3_integrated)' marker genes to file.")
gzout <- gzfile(
  description = here(
    "output",
    "marker_genes",
    "whole_cell",
    "uniquely_up",
    "integrated",
    paste0("cluster_3_integrated",
           ".uniquely_up.csv.gz")),
  open = "wb")

write.table(
  x = BDILP3_uniquely_up_noiseR_sig_sort_uniq_selected %>%
    as.data.frame() %>%
    tibble::rownames_to_column("gene_ID"),
  file = gzout,
  sep = ",",
  quote = FALSE,
  row.names = FALSE,
  col.names = TRUE)
close(gzout)

# top only + gene-of-interest
best_set <- BDILP3_uniquely_up_noiseR_sig_sort_uniq_selected[1:50, ]
```

```{r heat-uniquely-up-logExp-cluster-3-integrated, fig.cap = "Heatmap of log-expression values in each sample for the top uniquely upregulated marker genes. Each column is a sample, each row a gene. Colours are capped at -3 and 3 to preserve dynamic range.", fig.asp = 1.2, layout="l-body-outset"}
# heatmap
plotHeatmap(
  sce,
  features = rownames(best_set),
  columns = order(
    sce$cluster,
    sce$stage,
    sce$tissue,
    sce$donor,
    sce$group,
    sce$plate_number),
  colour_columns_by = c(
    "cluster",
    "stage",
    "tissue",
    "donor",
    "group",
    "plate_number"),
  cluster_cols = FALSE,
  center = TRUE,
  symmetric = TRUE,
  zlim = c(-3, 3),
  show_colnames = FALSE,
  annotation_row = data.frame(
    # TODO: temp trick to deal with the row-colouring problem
    cluster3.vs.1.2.4 = factor(ifelse(rownames(best_set) %in% rownames(B_uniquely_up_noiseR_sig), "DE", "not DE"), levels = c("DE")),
    cluster3.vs.1.2 = factor(ifelse(rownames(best_set) %in% rownames(D_uniquely_up_noiseR_sig), "DE", "not DE"), levels = c("DE")),
    cluster3.vs.4 = factor(ifelse(rownames(best_set) %in% rownames(I_uniquely_up_noiseR_sig), "DE", "not DE"), levels = c("DE")),
    cluster3.vs.2 = factor(ifelse(rownames(best_set) %in% rownames(L_uniquely_up_noiseR_sig), "DE", "not DE"), levels = c("DE")),
    cluster3.vs.1 = factor(ifelse(rownames(best_set) %in% rownames(P_uniquely_up_noiseR_sig), "DE", "not DE"), levels = c("DE")),
    cluster3.vs.1.vs.2.vs.4 = factor(ifelse(rownames(best_set) %in% rownames(cluster3_uniquely_up_noiseR_sig), "DE", "not DE"), levels = c("DE")),
    row.names = rownames(best_set)),
  main = paste0("Cluster: ", chosen, " ", x),
  column_annotation_colors = list(
    # Sig = c("Yes" = "red", "No" = "lightgrey"),
    cluster = cluster_colours,
    stage = stage_colours,
    tissue = tissue_colours,
    donor = donor_colours,
    group = group_colours,
    plate_number = plate_number_colours),
  color = hcl.colors(101, "Blue-Red 3"),
  fontsize = 7)
```

**Summary**: Here is a heatmap summarizes the row-normalized expression of the top 50 unique DE ordered by the ranking of FDR for each of the pairwise comparisons mentioned for cluster 3.
On top, is the classification of cells based on key experimental factor, such as cluster, stage, and tissue.
Whilst on the left, the colour bar indicates if a gene is a unique DE in a particular comparison, say, the POTEF gene is a unique DE only when comparing 3 to 4, but not the other 5 comparisons.

To interpret this heatmap, we can approach in several different ways.
Say, we can start by looking at the genes that are "DE in all 6 comparisons".

For instance, LEF1 and TCF is a known pair of transcription factors modulating T cell development in thymocytes. [Immunity 37, 813826, November 16, 2012], and SOX4 is a canonical HMG-TF working with miR181 to facilitate thymocyte differentiation. [Eur J Immunol. 1997 May;27(5):1292-5; J Exp Med (2018) 215 (11): 28872900]. 
Also, knockdown in CD3D expression is proven to affect the differentiation of alpha-beta and gamma-delta T cells in human [J Clin Invest. 2011;121(10):3872-3876], whilst TOX is a key factor involved in T cell lineage commitment in thymocytes [Nature Immunology volume 3, pages 272280 (2002)].

Another way round, we can also focus on genes that appear to be DE in a particular comparison only. 
For example, TRDC is the T cell receptor delta chain and is higher in expression in cluster 3 (i.e. mostly S3) when compared to cluster group of 1, 2, and 4 (i.e. mostly S1 and S2), whilst NPIPB gene family members known to modulate type I interferon [J. Microbiol. Immunol. Infect. 50, 277285 (2015)] were constantly expressed in both cluster 1 and 4 cells, but was slightly higher in expression in cluster 3 when compared to cluster 2.

:::

::: {.panel}

#### `Cluster_4_integrated` {.panel-name}

```{r}
############
# cluster 4
############

chosen <- "4"

# add description for the chosen cluster-group
x <- "(mostly Thymus.S3)"

# retain only significant markers (FDR<0.05) + keep only required output columns
F_uniquely_up_noiseR_sig <- F_uniquely_up_noiseR[F_uniquely_up_noiseR$FDR<0.05,][ ,c(1:3)]
J_uniquely_up_noiseR_sig <- J_uniquely_up_noiseR[J_uniquely_up_noiseR$FDR<0.05,][ ,c(1:3)]
N_uniquely_up_noiseR_sig <- N_uniquely_up_noiseR[N_uniquely_up_noiseR$FDR<0.05,][ ,c(1:3)]
R_uniquely_up_noiseR_sig <- R_uniquely_up_noiseR[R_uniquely_up_noiseR$FDR<0.05,][ ,c(1:3)]
cluster4_uniquely_up_noiseR_sig <- cluster4_uniquely_up_noiseR[cluster4_uniquely_up_noiseR$FDR<0.05,][ ,c(1:3)]

# add top column
F_uniquely_up_noiseR_sig$top <- 1:nrow(F_uniquely_up_noiseR_sig)
J_uniquely_up_noiseR_sig$top <- 1:nrow(J_uniquely_up_noiseR_sig)
N_uniquely_up_noiseR_sig$top <- 1:nrow(N_uniquely_up_noiseR_sig)
R_uniquely_up_noiseR_sig$top <- 1:nrow(R_uniquely_up_noiseR_sig)
cluster4_uniquely_up_noiseR_sig$top <- 1:nrow(cluster4_uniquely_up_noiseR_sig)

# unify S4 objects, sort by top (ascending) then FDR (ascending), keep only first unique entry for each marker
FJNR4_uniquely_up_noiseR_sig <- rbind2(F_uniquely_up_noiseR_sig,
                                       J_uniquely_up_noiseR_sig,
                                       N_uniquely_up_noiseR_sig,
                                       R_uniquely_up_noiseR_sig,
                                       cluster4_uniquely_up_noiseR_sig)

FJNR4_uniquely_up_noiseR_sig_sort <- FJNR4_uniquely_up_noiseR_sig[with(FJNR4_uniquely_up_noiseR_sig, order(top, FDR)), ]
FJNR4_uniquely_up_noiseR_sig_sort_uniq <- FJNR4_uniquely_up_noiseR_sig_sort[unique(rownames(FJNR4_uniquely_up_noiseR_sig_sort)), ]

# # de-select unannotated/ not well-characterised genes
# deselected <- c("MTRNR1L12", "MTRNR2L8", "NPIPB12", "NPIPB4", "NPIPB3", "NPIPB13", "NPIPB11", "NPIPB15", "MALAT1")
deselected <- c()
FJNR4_uniquely_up_noiseR_sig_sort_uniq_selected <- FJNR4_uniquely_up_noiseR_sig_sort_uniq[!(rownames(FJNR4_uniquely_up_noiseR_sig_sort_uniq) %in% deselected), ]

# export DGE lists
saveRDS(
  FJNR4_uniquely_up_noiseR_sig_sort_uniq_selected,
  here("data", "marker_genes", "whole_cell", "C094_Pellicci.uniquely_up.cluster_4_integrated.rds"),
  compress = "xz")

dir.create(here("output", "marker_genes", "whole_cell", "uniquely_up", "integrated"), recursive = TRUE)

message("Writing 'uniquely_up (cluster_4_integrated)' marker genes to file.")
gzout <- gzfile(
  description = here(
    "output",
    "marker_genes",
    "whole_cell",
    "uniquely_up",
    "integrated",
    paste0("cluster_4_integrated",
           ".uniquely_up.csv.gz")),
  open = "wb")

write.table(
  x = FJNR4_uniquely_up_noiseR_sig_sort_uniq_selected %>%
    as.data.frame() %>%
    tibble::rownames_to_column("gene_ID"),
  file = gzout,
  sep = ",",
  quote = FALSE,
  row.names = FALSE,
  col.names = TRUE)
close(gzout)

# top only + gene-of-interest
best_set <- FJNR4_uniquely_up_noiseR_sig_sort_uniq_selected[1:18, ]
```

```{r heat-uniquely-up-logExp-cluster-4-integrated, fig.cap = "Heatmap of log-expression values in each sample for the top uniquely upregulated marker genes. Each column is a sample, each row a gene. Colours are capped at -3 and 3 to preserve dynamic range.", fig.asp = 1.2, layout="l-body-outset"}
# heatmap
plotHeatmap(
  sce,
  features = rownames(best_set),
  columns = order(
    sce$cluster,
    sce$stage,
    sce$tissue,
    sce$donor,
    sce$group,
    sce$plate_number),
  colour_columns_by = c(
    "cluster",
    "stage",
    "tissue",
    "donor",
    "group",
    "plate_number"),
  cluster_cols = FALSE,
  center = TRUE,
  symmetric = TRUE,
  zlim = c(-3, 3),
  show_colnames = FALSE,
  annotation_row = data.frame(
    # TODO: temp trick to deal with the row-colouring problem
    cluster4.vs.1.2 = factor(ifelse(rownames(best_set) %in% rownames(F_uniquely_up_noiseR_sig), "DE", "not DE"), levels = c("DE")),
    cluster4.vs.3 = factor(ifelse(rownames(best_set) %in% rownames(J_uniquely_up_noiseR_sig), "DE", "not DE"), levels = c("DE")),
    cluster4.vs.2 = factor(ifelse(rownames(best_set) %in% rownames(N_uniquely_up_noiseR_sig), "DE", "not DE"), levels = c("DE")),
    cluster4.vs.1 = factor(ifelse(rownames(best_set) %in% rownames(R_uniquely_up_noiseR_sig), "DE", "not DE"), levels = c("DE")),
    cluster4.vs.1.vs.2.vs.3 = factor(ifelse(rownames(best_set) %in% rownames(cluster4_uniquely_up_noiseR_sig), "DE", "not DE"), levels = c("DE")),
    row.names = rownames(best_set)),
  main = paste0("Cluster: ", chosen, " ", x),
  column_annotation_colors = list(
    # Sig = c("Yes" = "red", "No" = "lightgrey"),
    cluster = cluster_colours,
    stage = stage_colours,
    tissue = tissue_colours,
    donor = donor_colours,
    group = group_colours,
    plate_number = plate_number_colours),
  color = hcl.colors(101, "Blue-Red 3"),
  fontsize = 7)
```

**Summary**: Cluster `4` did not have as many DE when pairwise-compared to the other clusters, except when compared to cluster `3`.
This may suggest that cluster `4` could represents an "interim development stage" that shared lots of transcriptomic fingerprints with the other clusters.
Indeed, this goes in line with the fact that cluster `4` is enriched by cells from Thymic S3, which supposed to be formed from `Thymic S2` (enriched in cluster `3`) and will be developed into `Blood S3` (that found mostly in cluster `1`).

In terms of marker genes, the only clear marker shown in this cluster is `MALAT1` (which is a long non-coding RNA highly and commonly variably expressed in different T cells; it also is known to be non-essential to the T cell development [RNA Biol. 2018; 15(12): 14771486)]).
But notably, this cluster did have higher expression level of `KLRB1` than cluster `3`, indicating its `S3` identity. Moreover, this cluster also possesses marker such as `SYNE2` that is known to be expressed in gamma-delta T cells [Nat Immunol 13, 10001009 (2012)], supporting this cluster contains cells relevant to this study rather than being contaminant.

Further, based on the fact that most of the cells of cluster `4` is originated from donor `2` and two of the markers, `MTRNR2L12` and `MTRNR2L8`, of cluster 4 are mitochondrial transcripts, this makes us wonder whether it could indicate a technical error and/or stress-induced mitochondrial expression sourced from samples of donor `2`.
However, based on the QC plot broken down by `cluster` and `donor` (Figure \@ref(fig:qcplot-by-cluster-donor), we cannot spot any apparent difference between donor `2` samples and the rest in terms of both library size or mitochondrial transcript proportion, thus we rule out this hypothesis.

An additional point-to-note is that we observe a number of `NPIPB`-related genes appears as markers for this cluster.
`NPIPB` is supposed to a gene family with members of highly similar sequence.
As we only perform 3'-tag sequencing and reads from any member of this gene family could potentially mis-aligned during the mapping step, which could in turn potentially lead to the variation of the gene expression we observed for cluster `4`.
With this regard, we will need to be cautious and try not to make some critical statement or build any key hypothesis based on the NPIPB-related genes.

```{r qcplot-by-cluster-donor, fig.cap = "Distributions of various QC metrics for all single cells in the dataset, stratified by cluster and donor This includes the library sizes (log10), number of expressed genes (log10), and proportion of reads mapped to spike-in transcripts or mitochondrial genes.", layout = "l-page", fig.asp = 0.7}
plot_grid(
  
  ggplot(
    data = as.data.frame(colData(sce)),
    aes(x = cluster, y = sum, colour = donor, fill = donor)) +
    geom_violin(scale = "width", width = 0.8, alpha = 0.2) +
    theme_cowplot(font_size = 7) +
    scale_colour_manual(values = donor_colours) + 
    scale_fill_manual(values = donor_colours) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_y_log10(limits = c(1, NA)) +
    annotation_logticks(
      sides = "l",
      short = unit(0.03, "cm"),
      mid = unit(0.06, "cm"),
      long = unit(0.09, "cm")),
  
  ggplot(
    data = as.data.frame(colData(sce)),
    aes(x = cluster, y = altexps_ERCC_percent, colour = donor, fill = donor)) +
    geom_violin(scale = "width", width = 0.8, alpha = 0.2) +
    ylim(0, 100) +
    theme_cowplot(font_size = 7) +
    scale_colour_manual(values = donor_colours) + 
    scale_fill_manual(values = donor_colours) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)),
  
  ggplot(
    data = as.data.frame(colData(sce)),
    aes(x = cluster, y = detected, colour = donor, fill = donor)) +
    geom_violin(scale = "width", width = 0.8, alpha = 0.2) +
    theme_cowplot(font_size = 7) +
    scale_colour_manual(values = donor_colours) + 
    scale_fill_manual(values = donor_colours) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_y_log10(limits = c(1, NA)) +
    annotation_logticks(
      sides = "l",
      short = unit(0.03, "cm"),
      mid = unit(0.06, "cm"),
      long = unit(0.09, "cm")),
  
  ggplot(
    data = as.data.frame(colData(sce)),
    aes(x = cluster, y = subsets_Mito_percent, colour = donor, fill = donor)) +
    geom_violin(scale = "width", width = 0.8, alpha = 0.2) +
    ylim(0, 100) +
    theme_cowplot(font_size = 7) +
    scale_colour_manual(values = donor_colours) + 
    scale_fill_manual(values = donor_colours) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)),
  
  ncol = 2)
```

```{r qcplot-by-cluster-donor-linear, fig.cap = "Distributions of various QC metrics for all single cells in the dataset, stratified by cluster and donor This includes the library sizes (linear), number of expressed genes (linear), and proportion of reads mapped to spike-in transcripts or mitochondrial genes.", layout = "l-page", fig.asp = 0.7}
plot_grid(
  
  ggplot(
    data = as.data.frame(colData(sce)),
    aes(x = cluster, y = sum, colour = donor, fill = donor)) +
    geom_violin(scale = "width", width = 0.8, alpha = 0.2) +
    theme_cowplot(font_size = 7) +
    scale_colour_manual(values = donor_colours) + 
    scale_fill_manual(values = donor_colours) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)),
  
  ggplot(
    data = as.data.frame(colData(sce)),
    aes(x = cluster, y = altexps_ERCC_percent, colour = donor, fill = donor)) +
    geom_violin(scale = "width", width = 0.8, alpha = 0.2) +
    ylim(0, 100) +
    theme_cowplot(font_size = 7) +
    scale_colour_manual(values = donor_colours) + 
    scale_fill_manual(values = donor_colours) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)),
  
  ggplot(
    data = as.data.frame(colData(sce)),
    aes(x = cluster, y = detected, colour = donor, fill = donor)) +
    geom_violin(scale = "width", width = 0.8, alpha = 0.2) +
    theme_cowplot(font_size = 7) +
    scale_colour_manual(values = donor_colours) + 
    scale_fill_manual(values = donor_colours) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)),
  
  ggplot(
    data = as.data.frame(colData(sce)),
    aes(x = cluster, y = subsets_Mito_percent, colour = donor, fill = donor)) +
    geom_violin(scale = "width", width = 0.8, alpha = 0.2) +
    ylim(0, 100) +
    theme_cowplot(font_size = 7) +
    scale_colour_manual(values = donor_colours) + 
    scale_fill_manual(values = donor_colours) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)),
  
  ncol = 2)
```

:::

::: {.panel}

#### `Cluster_2_integrated` {.panel-name}

```{r}
############
# cluster 2
############

chosen <- "2"

# add description for the chosen cluster-group
x <- "(S3-mix more Thymus)"

# retain only significant markers (FDR<0.05) + keep only required output columns
K_uniquely_up_noiseR_sig <- K_uniquely_up_noiseR[K_uniquely_up_noiseR$FDR<0.05,][ ,c(1:3)]
M_uniquely_up_noiseR_sig <- M_uniquely_up_noiseR[M_uniquely_up_noiseR$FDR<0.05,][ ,c(1:3)]
H_uniquely_up_noiseR_sig <- H_uniquely_up_noiseR[H_uniquely_up_noiseR$FDR<0.05,][ ,c(1:3)]
cluster2_uniquely_up_noiseR_sig <- cluster2_uniquely_up_noiseR[cluster2_uniquely_up_noiseR$FDR<0.05,][ ,c(1:3)]

# add top column
K_uniquely_up_noiseR_sig$top <- 1:nrow(K_uniquely_up_noiseR_sig)
M_uniquely_up_noiseR_sig$top <- 1:nrow(M_uniquely_up_noiseR_sig)
# TODO: fix error - exclude list from rbind2 if empty
# H_uniquely_up_noiseR_sig$top <- 1:nrow(H_uniquely_up_noiseR_sig)
# cluster2_uniquely_up_noiseR_sig$top <- 1:nrow(cluster2_uniquely_up_noiseR_sig)

# unify S4 objects, sort by top (ascending) then FDR (ascending), keep only first unique entry for each marker
KMH2_uniquely_up_noiseR_sig <- rbind2(K_uniquely_up_noiseR_sig,
                                      M_uniquely_up_noiseR_sig,
                                      H_uniquely_up_noiseR_sig,
                                      cluster2_uniquely_up_noiseR_sig)

KMH2_uniquely_up_noiseR_sig_sort <- KMH2_uniquely_up_noiseR_sig[with(KMH2_uniquely_up_noiseR_sig, order(top, FDR)), ]
KMH2_uniquely_up_noiseR_sig_sort_uniq <- KMH2_uniquely_up_noiseR_sig_sort[unique(rownames(KMH2_uniquely_up_noiseR_sig_sort)), ]

# # de-select unannotated/ not well-characterised genes
deselected <- c()
KMH2_uniquely_up_noiseR_sig_sort_uniq_selected <- KMH2_uniquely_up_noiseR_sig_sort_uniq[!(rownames(KMH2_uniquely_up_noiseR_sig_sort_uniq) %in% deselected), ]

# export DGE lists
saveRDS(
  KMH2_uniquely_up_noiseR_sig_sort_uniq_selected,
  here("data", "marker_genes", "whole_cell", "C094_Pellicci.uniquely_up.cluster_2_integrated.rds"),
  compress = "xz")

dir.create(here("output", "marker_genes", "whole_cell", "uniquely_up", "integrated"), recursive = TRUE)

message("Writing 'uniquely_up (cluster_2_integrated)' marker genes to file.")
gzout <- gzfile(
  description = here(
    "output",
    "marker_genes",
    "whole_cell",
    "uniquely_up",
    "integrated",
    paste0("cluster_2_integrated",
           ".uniquely_up.csv.gz")),
  open = "wb")

write.table(
  x = KMH2_uniquely_up_noiseR_sig_sort_uniq_selected %>%
    as.data.frame() %>%
    tibble::rownames_to_column("gene_ID"),
  file = gzout,
  sep = ",",
  quote = FALSE,
  row.names = FALSE,
  col.names = TRUE)
close(gzout)

# top only + gene-of-interest
best_set <- KMH2_uniquely_up_noiseR_sig_sort_uniq_selected[1:50, ]
```

```{r heat-uniquely-up-logExp-cluster-2-integrated, fig.cap = "Heatmap of log-expression values in each sample for the top uniquely upregulated marker genes. Each column is a sample, each row a gene. Colours are capped at -3 and 3 to preserve dynamic range.", fig.asp = 1.2, layout="l-body-outset"}
# heatmap
plotHeatmap(
  sce,
  features = rownames(best_set),
  columns = order(
    sce$cluster,
    sce$stage,
    sce$tissue,
    sce$donor,
    sce$group,
    sce$plate_number),
  colour_columns_by = c(
    "cluster",
    "stage",
    "tissue",
    "donor",
    "group",
    "plate_number"),
  cluster_cols = FALSE,
  center = TRUE,
  symmetric = TRUE,
  zlim = c(-3, 3),
  show_colnames = FALSE,
  annotation_row = data.frame(
    # TODO: temp trick to deal with the row-colouring problem
    cluster2.vs.3 = factor(ifelse(rownames(best_set) %in% rownames(K_uniquely_up_noiseR_sig), "DE", "not DE"), levels = c("DE")),
    cluster2.vs.4 = factor(ifelse(rownames(best_set) %in% rownames(M_uniquely_up_noiseR_sig), "DE", "not DE"), levels = c("DE")),
    # TODO: work out how to remove `fill` from `not_DE`
    cluster2.vs.1 = factor(ifelse(rownames(best_set) %in% rownames(H_uniquely_up_noiseR_sig), "DE", "not DE"), levels = c("DE", "not DE")),
    cluster2.vs.1.vs.3.vs.4 = factor(ifelse(rownames(best_set) %in% rownames(cluster2_uniquely_up_noiseR_sig), "DE", "not DE"), levels = c("DE", "not DE")),
    row.names = rownames(best_set)),
  main = paste0("Cluster: ", chosen, " ", x),
  column_annotation_colors = list(
    # Sig = c("Yes" = "red", "No" = "lightgrey"),
    cluster = cluster_colours,
    stage = stage_colours,
    tissue = tissue_colours,
    donor = donor_colours,
    group = group_colours,
    plate_number = plate_number_colours),
  color = hcl.colors(101, "Blue-Red 3"),
  fontsize = 7)
```

**Summary**: see `Cluster_1_integrated` summary about comment on this cluster.

:::

::: {.panel}

#### `Cluster_1_integrated` {.panel-name}

```{r}
############
# cluster 1
############

chosen <- "1"

# add description for the chosen cluster-group
x <- "(S3-mix more Blood)"

# retain only significant markers (FDR<0.05) + keep only required output columns
G_uniquely_up_noiseR_sig <- G_uniquely_up_noiseR[G_uniquely_up_noiseR$FDR<0.05,][ ,c(1:3)]
O_uniquely_up_noiseR_sig <- O_uniquely_up_noiseR[O_uniquely_up_noiseR$FDR<0.05,][ ,c(1:3)]
Q_uniquely_up_noiseR_sig <- Q_uniquely_up_noiseR[Q_uniquely_up_noiseR$FDR<0.05,][ ,c(1:3)]
cluster1_uniquely_up_noiseR_sig <- cluster1_uniquely_up_noiseR[cluster1_uniquely_up_noiseR$FDR<0.05,][ ,c(1:3)]

# add top column
G_uniquely_up_noiseR_sig$top <- 1:nrow(G_uniquely_up_noiseR_sig)
O_uniquely_up_noiseR_sig$top <- 1:nrow(O_uniquely_up_noiseR_sig)
Q_uniquely_up_noiseR_sig$top <- 1:nrow(Q_uniquely_up_noiseR_sig)
cluster1_uniquely_up_noiseR_sig$top <- 1:nrow(cluster1_uniquely_up_noiseR_sig)

# unify S4 objects, sort by top (ascending) then FDR (ascending), keep only first unique entry for each marker
GOQ1_uniquely_up_noiseR_sig <- rbind2(G_uniquely_up_noiseR_sig,
                                      O_uniquely_up_noiseR_sig,
                                      Q_uniquely_up_noiseR_sig,
                                      cluster1_uniquely_up_noiseR_sig)

GOQ1_uniquely_up_noiseR_sig_sort <- GOQ1_uniquely_up_noiseR_sig[with(GOQ1_uniquely_up_noiseR_sig, order(top, FDR)), ]
GOQ1_uniquely_up_noiseR_sig_sort_uniq <- GOQ1_uniquely_up_noiseR_sig_sort[unique(rownames(GOQ1_uniquely_up_noiseR_sig_sort)), ]

# # de-select unannotated/ not well-characterised genes
deselected <- c()
GOQ1_uniquely_up_noiseR_sig_sort_uniq_selected <- GOQ1_uniquely_up_noiseR_sig_sort_uniq[!(rownames(GOQ1_uniquely_up_noiseR_sig_sort_uniq) %in% deselected), ]

# export DGE lists
saveRDS(
  GOQ1_uniquely_up_noiseR_sig_sort_uniq_selected,
  here("data", "marker_genes", "whole_cell", "C094_Pellicci.uniquely_up.cluster_1_integrated.rds"),
  compress = "xz")

dir.create(here("output", "marker_genes", "whole_cell", "uniquely_up", "integrated"), recursive = TRUE)

message("Writing 'uniquely_up (cluster_1_integrated)' marker genes to file.")
gzout <- gzfile(
  description = here(
    "output",
    "marker_genes",
    "whole_cell",
    "uniquely_up",
    "integrated",
    paste0("cluster_1_integrated",
           ".uniquely_up.csv.gz")),
  open = "wb")

write.table(
  x = GOQ1_uniquely_up_noiseR_sig_sort_uniq_selected %>%
    as.data.frame() %>%
    tibble::rownames_to_column("gene_ID"),
  file = gzout,
  sep = ",",
  quote = FALSE,
  row.names = FALSE,
  col.names = TRUE)
close(gzout)

# top only + gene-of-interest
best_set <- GOQ1_uniquely_up_noiseR_sig_sort_uniq_selected[1:50, ]
```

```{r heat-uniquely-up-logExp-cluster-1-integrated, fig.cap = "Heatmap of log-expression values in each sample for the top uniquely upregulated marker genes. Each column is a sample, each row a gene. Colours are capped at -3 and 3 to preserve dynamic range.", fig.asp = 1.2, layout="l-body-outset"}
# heatmap
plotHeatmap(
  sce,
  features = rownames(best_set),
  columns = order(
    sce$cluster,
    sce$stage,
    sce$tissue,
    sce$donor,
    sce$group,
    sce$plate_number),
  colour_columns_by = c(
    "cluster",
    "stage",
    "tissue",
    "donor",
    "group",
    "plate_number"),
  cluster_cols = FALSE,
  center = TRUE,
  symmetric = TRUE,
  zlim = c(-3, 3),
  show_colnames = FALSE,
  annotation_row = data.frame(
    # TODO: temp trick to deal with the row-colouring problem
    cluster1.vs.2 = factor(ifelse(rownames(best_set) %in% rownames(G_uniquely_up_noiseR_sig), "DE", "not DE"), levels = c("DE")),
    cluster1.vs.3 = factor(ifelse(rownames(best_set) %in% rownames(O_uniquely_up_noiseR_sig), "DE", "not DE"), levels = c("DE")),
    cluster1.vs.4 = factor(ifelse(rownames(best_set) %in% rownames(Q_uniquely_up_noiseR_sig), "DE", "not DE"), levels = c("DE")),
    cluster1.vs.2.vs.3.vs.4 = factor(ifelse(rownames(best_set) %in% rownames(cluster1_uniquely_up_noiseR_sig), "DE", "not DE"), levels = c("DE")),
    row.names = rownames(best_set)),
  main = paste0("Cluster: ", chosen, " ", x),
  column_annotation_colors = list(
    # Sig = c("Yes" = "red", "No" = "lightgrey"),
    cluster = cluster_colours,
    stage = stage_colours,
    tissue = tissue_colours,
    donor = donor_colours,
    group = group_colours,
    plate_number = plate_number_colours),
  color = hcl.colors(101, "Blue-Red 3"),
  fontsize = 7)
```

**Summary**: The next two clusters I would like to talk about is cluster 1 and 2.
Cell-types/ -stages of cluster `1` and cluster `2` are highly similar as implied by their UMAP distribution (Figure \@ref(fig:clusterplot-umap)). 
When looked into the heatmap for both cluster `1` and `2`, the first thing that draw my particular attention is that both cluster 1 and 2 contains number of interferon- and interleukin-related genes (such as members of the Interferon Induced Transmembrane IFITM family, and of the interleukin-receptor family) and also the class I HLA genes (including HLA-A, B, C, E and F).

Having said that these two clusters shared a number of similarities, they do show differences. 
In terms of `tissue` and `stage`, both cluster `1` and `2` are enriched by `S3` cells and the only recognizable difference between the two clusters is that, cluster `2` seems to be `S3`-mix contains more `thymus` cells, whilst cluster 1 is a `S3`-mix as well but with more cells from `blood.`
In terms of DE, cluster `1` still contain a handful of markers highly expressed in cluster 1, but not in cluster `2`. 
This includes: CCL5, i.e. a chemokine that known to be involved in thymocyte trafficking [Oncotarget. 2017; 8:83-94] and capable to alter the coordinated maturation and migration of thymocytes [Cell Adhesion & Migration 2010 4:3, 382-390].
EFHD2, a calcium binding adapter protein required for human T cellmediated cytotoxicity [The Journal of Immunology, 2018, 201].
GNLY is a component of cytotoxic granules that cytotoxic granules that can cause lipid degradation secreted by NK and T cells [Innate Immunity Vol 24, Issue 8, 2018, pp452-465].
Last but not least is NKG7, which is a factor essential for cytotoxic degranulation and activation of proinflammatory responses triggered by T cells [Nature Immunology volume 21, pages 11391140 (2020)].

:::

::::::

DGE lists of these comparisons are available in [`output/marker_genes/whole_cell/uniquely_up/integrated/`](..output/marker_genes/whole_cell/uniquely_up/integrated/).

## Supplemental info

**Comment**: We have no idea how well the CELSeq protocol could capture the expression of the T-cell-receptor component and the suggested known markers, i.e. TRDV2, TRGV9, may not be a good candidate for defining whether a cluster contains cells relevant to gamma-delta T cells.

```{r}
############################################
# Sanity check (TRDV2 and TRGV9 expression)

assay(sce, "log2cpm") <- edgeR::cpm(
  counts(sce),
  log = TRUE,
  lib.size = colSums(counts(sce)))
```

```{r violin-trdv2-trgv9, fig.asp = 1, layout="l-body-outset"}
plot_grid(
  
  plotExpression(
    sce,
    features = "TRDV2",
    x = "cluster",
    exprs_values = "log2cpm",
    colour_by = "cluster",
    other_fields = "stage") +
    ggtitle("Expression of TRDV2") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_colour_manual(values = cluster_colours, name = "cluster") +
    facet_grid(~stage),
  
  plotExpression(
    sce,
    features = "TRGV9",
    x = "cluster",
    exprs_values = "log2cpm",
    colour_by = "cluster",
    other_fields = "stage") +
    ggtitle("Expression of TRGV9") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_colour_manual(values = cluster_colours, name = "cluster") +
    facet_grid(~stage),
  
  ncol= 1,
  align ="h"
)
```

# Session info {.appendix}

<summary>The analysis and this document were prepared using the following software (click triangle to expand)</summary>
<details>

```{r}
sessioninfo::session_info()
```

</details>
